---
title: "GSS Variable Recoding and Preparation (2016–2024)"
editor: visual
---

## Overview

Recodes subjective items from the 2016–2024 General Social Survey cumulative file into binary (`c_`-prefixed) outcome variables and saves the result as an RDS for downstream analysis. Three operations are applied per variable:

1.  **Coalescing** — merges split-ballot variants (Form 1/2, `-G`/`-X`/`-Y`/`-V`/`-NV` experiments) into a single variable per construct.
2.  **Binarization** — converts ordinal scales into 0/1 outcomes using distribution-based split rules.
3.  **Filtering** — restricts to 2016, 2018, 2022, and 2024.

Only variables in the analytic core (`data/gss24_attitudes_opinions.csv`, `analytic_core == 1`) are used in downstream files; recodes for excluded items are omitted here.

------------------------------------------------------------------------

## 1. Setup and Data Loading

```{r setup, message=FALSE, warning=FALSE}
source("00_setup.R")

if (file.exists(GSS_DTA_PATH)) {
  gss_data <- read_dta(GSS_DTA_PATH)
} else {
  stop("Data file not found: ", GSS_DTA_PATH)
}
```

------------------------------------------------------------------------

## 2. Recoding Functions

```{r functions}
#' Replicates Stata's `recode` command for binary variables.
gssrecode <- function(data, vars, to0, to1) {
  for (var in vars) {
    if (!var %in% names(data)) {
      warning(paste0("Variable '", var, "' not found. Skipping."))
      next
    }
    new_var_name <- paste0("c_", var)
    data <- data %>%
      mutate(!!new_var_name := case_when(
        .data[[var]] %in% to0 ~ 0,
        .data[[var]] %in% to1 ~ 1,
        TRUE ~ NA_real_
      ))
  }
  return(data)
}

#' Recodes 3-category variables into a binary split based on distribution.
gss3cats <- function(data, vars) {
  for (var in vars) {
    if (!var %in% names(data)) next
    num1 <- sum(data[[var]] == 1, na.rm = TRUE)
    num3 <- sum(data[[var]] == 3, na.rm = TRUE)
    if (num1 >= num3) {
      data <- gssrecode(data, var, to0 = c(2, 3), to1 = 1)
    } else {
      data <- gssrecode(data, var, to0 = 3, to1 = c(1, 2))
    }
  }
  return(data)
}

#' Recodes 5-category variables into a binary split based on distribution.
gss5cats <- function(data, vars) {
  for (var in vars) {
    if (!var %in% names(data)) next
    num12 <- sum(data[[var]] %in% c(1, 2), na.rm = TRUE)
    num45 <- sum(data[[var]] %in% c(4, 5), na.rm = TRUE)
    if (num12 >= num45) {
      data <- gssrecode(data, var, to0 = c(3, 4, 5), to1 = c(1, 2))
    } else {
      data <- gssrecode(data, var, to0 = c(4, 5), to1 = c(1, 2, 3))
    }
  }
  return(data)
}

#' Recodes 7-category variables into a binary split based on distribution.
gss7cats <- function(data, vars) {
  for (var in vars) {
    if (!var %in% names(data)) next
    num123 <- sum(data[[var]] %in% c(1, 2, 3), na.rm = TRUE)
    num567 <- sum(data[[var]] %in% c(5, 6, 7), na.rm = TRUE)
    if (num123 >= num567) {
      data <- gssrecode(data, var, to0 = c(4, 5, 6, 7), to1 = c(1, 2, 3))
    } else {
      data <- gssrecode(data, var, to0 = c(5, 6, 7), to1 = c(1, 2, 3, 4))
    }
  }
  return(data)
}
```

------------------------------------------------------------------------

## 3. Recoding

::: {.callout-note title="Experiment types in this file"}
**`-G` (grid):** Web and phone, FORM=1. Web respondents see items in a condensed grid layout; phone respondents assigned FORM=1 are also recorded in `-G` variables despite not seeing a grid (codebook p. 30). Original non-grid item on FORM=2. Effects may be selective across items.

**`-Y` (wording):** Both modes. Two distinct subtypes:

-   *Traditional* — alternate question wording or intro on Form 2; original on Form 1. (`NAT*` series, `CONLABORY`, `MARSAMEY`, `XMOVIEY`.) Most gaps negligible; `NATCITY`/`NATFARE` are exceptions (see below).
-   *Gender-neutral* — Form 2 replaces gendered terms (e.g., "policeman") with gender-neutral equivalents (e.g., "police officer"); original wording on Form 1. (`SPKATH/Y`, `LIBATH/Y`, `COLCOM/Y`, `LETDIE1/Y`, `POLHITOK/Y`, etc.) Negligible gaps. Note: `RACOPEN` was included in *both* the gender-neutral and volunteered response experiments; its variants are `RACOPENV`/`RACOPENNV`, not `RACOPENY` (codebook p. 31).

**`-X` (wording):** Both modes. Extended alternate intro wording on Form 1; original on Form 2. Used for `RACDIF` series — the Form 1 intro informs respondents that the questions have been asked for almost 50 years. Negligible gaps.

**`-V`/`-NV` (volunteered response):** Web mode only (MODE=4). `-V` (FORM=2) shows the volunteered response as an on-screen option; `-NV` (FORM=1) omits it entirely. FTF/phone respondents retain the ability to volunteer responses regardless of form; their data appear in the unmodified base variable. Large V vs. NV gaps are expected by design. The codebook warns that merging `-V`, `-NV`, and unmodified variables produces anomalous trends relative to pre-2022 data; coalescing is valid here only because this paper applies mode-specific weights (`wtssps_web`, `wtssps_f2fp`) that account for mode-specific subsamples.
:::

### National Spending Items

Coalesces Form 2 variants (`-Y` suffix) into main variable names before recoding. `NATCITY` and `NATFARE` are excluded from this coalesce: the wording gap between forms is large and stable (\~0.15pp and \~0.20pp respectively across all years and modes), so both forms are treated as separate analytic variables.

```{r nat-spend}
gss_data <- gss_data %>%
  mutate(
    natspac  = if_else(form == 2 & !is.na(natspacy),  natspacy,  natspac),
    natenvir = if_else(form == 2 & !is.na(natenviy),  natenviy,  natenvir),
    natheal  = if_else(form == 2 & !is.na(nathealy),  nathealy,  natheal),
    natdrug  = if_else(form == 2 & !is.na(natdrugy),  natdrugy,  natdrug),
    nateduc  = if_else(form == 2 & !is.na(nateducy),  nateducy,  nateduc),
    natarms  = if_else(form == 2 & !is.na(natarmsy),  natarmsy,  natarms),
    natrace  = if_else(form == 2 & !is.na(natracey),  natracey,  natrace),
    nataid   = if_else(form == 2 & !is.na(nataidy),   nataidy,   nataid),
    natcrime = if_else(form == 2 & !is.na(natcrimy),  natcrimy,  natcrime)
  )

natspend_vars <- c("natspac", "natenvir", "natheal", "natcity", "natcrime",
                   "natdrug", "nateduc", "natrace", "natarms", "nataid",
                   "natfare", "natcityy", "natfarey", "natroad", "natsoc", "natchld", "natsci", "natenrgy")
gss_data <- gss3cats(gss_data, vars = natspend_vars)
```

### Confidence and Civil Liberties

Handles Form 1/2 wording experiments (traditional and gender-neutral) and the volunteered response (`-V`/`-NV`) experiment (web mode only).

```{r civil-liberties, warning=FALSE}
gss_data <- gss_data %>%
  mutate(
    # -Y traditional wording (Form 2 alternate; both modes; negligible gaps)
    conlabor = if_else(form == 2 & !is.na(conlabory), conlabory, conlabor),  # "labor unions" replacing "organized labor"

    # -Y gender-neutral wording (Form 2; both modes; negligible gaps — most replace "policeman" with "police officer"; letdie1 updates pronouns)
    colcom   = if_else(form == 2 & !is.na(colcomy),   colcomy,   colcom),
    spkath   = if_else(form == 2 & !is.na(spkathy),   spkathy,   spkath),
    spkrac   = if_else(form == 2 & !is.na(spkracy),   spkracy,   spkrac),
    spkcom   = if_else(form == 2 & !is.na(spkcomy),   spkcomy,   spkcom),
    spkmslm  = if_else(form == 2 & !is.na(spkmslmy),  spkmslmy,  spkmslm),
    libath   = if_else(form == 2 & !is.na(libathy),   libathy,   libath),
    librac   = if_else(form == 2 & !is.na(libracy),   libracy,   librac),
    libcom   = if_else(form == 2 & !is.na(libcomy),   libcomy,   libcom),
    libmslm  = if_else(form == 2 & !is.na(libmslmy),  libmslmy,  libmslm),
    letdie1  = if_else(form == 2 & !is.na(letdie1y),  letdie1y,  letdie1),   # pronoun update

    # -V/-NV volunteered response experiment (web only, MODE=4; large V-NV gaps expected by design)
    fepol    = if_else(!is.na(fepolv),   fepolv,   if_else(!is.na(fepolnv),   fepolnv,   fepol)),
    racopen  = if_else(!is.na(racopenv), racopenv, if_else(!is.na(racopennv), racopennv, racopen)),
    bible    = if_else(!is.na(biblev),   biblev,   if_else(!is.na(biblenv),   biblenv,   bible)),
    divlaw   = if_else(!is.na(divlawv),  divlawv,  if_else(!is.na(divlawnv),  divlawnv,  divlaw)),
    kidssol  = if_else(!is.na(kidssolv) & kidssolv <= 5, kidssolv,
                       if_else(!is.na(kidssolnv) & kidssolnv <= 5, kidssolnv, kidssol)),  # cap at 5 to exclude volunteered "no ideal"
    grass    = if_else(!is.na(grassv),    grassv,    if_else(!is.na(grassnv),    grassnv,    grass)),
    prayer   = if_else(!is.na(prayerv),   prayerv,   if_else(!is.na(prayernv),   prayernv,   prayer)),
    postlife = if_else(!is.na(postlifev), postlifev, if_else(!is.na(postlifenv), postlifenv, postlife)),
    fair     = if_else(!is.na(fairv),     fairv,     if_else(!is.na(fairnv),     fairnv,     fair)),
    helpful  = if_else(!is.na(helpfulv),  helpfulv,  if_else(!is.na(helpfulnv),  helpfulnv,  helpful)),
    trust    = if_else(!is.na(trustv),    trustv,    if_else(!is.na(trustnv),    trustnv,    trust)),
    getahead = if_else(!is.na(getaheadv), getaheadv, if_else(!is.na(getaheadnv), getaheadnv, getahead)),
    courts   = if_else(!is.na(courtsv),   courtsv,   if_else(!is.na(courtsnv),   courtsnv,   courts))    # volunteered: "about right"
  )

confidence_vars <- c("confinan", "conbus", "conclerg", "coneduc", "confed",
                     "conlabor", "conpress", "conmedic", "contv", "conjudge",
                     "consci", "conlegis", "conarmy")
gss_data <- gss3cats(gss_data, vars = confidence_vars)

gss_data <- gssrecode(gss_data, vars = c("colath", "colrac", "colcom", "colmslm"), to0 = 4, to1 = 5)
gss_data <- gssrecode(gss_data, vars = c("libath", "librac", "libcom", "libmslm", "fepol", "racopen"), to0 = 1, to1 = 2)
gss_data <- gssrecode(gss_data, vars = c("spkath", "spkrac", "spkcom", "spkmslm"), to0 = 1, to1 = 2)
```

### Abortion and Police

Abortion items coalesce `-G` variants (grid layout, FORM=1) into the base variable names. The codebook notes that grid effects may be selective across items; `ABANYG` in particular showed grid effects despite not appearing in a grid (codebook p. 30). Police items coalesce `-Y` gender-neutral variants (Form 2: "police officer" replacing "policeman"; Table 4).

```{r abortion-police}
abortion_vars <- c("abdefect", "abnomore", "abhlth", "abpoor", "abrape", "absingle", "abany")
for (var in abortion_vars) {
  source_var <- paste0(var, "g")
  gss_data <- gss_data %>%
    mutate(!!sym(var) := if_else(form == 1 & !is.na(.data[[source_var]]), .data[[source_var]], .data[[var]]))  # -G
}
gss_data <- gssrecode(gss_data, vars = abortion_vars, to0 = 2, to1 = 1)

police_vars <- c("polhitok", "polabuse", "polattak")
for (var in police_vars) {
  source_var <- paste0(var, "y")
  gss_data <- gss_data %>%
    mutate(!!sym(var) := if_else(form == 2 & !is.na(.data[[source_var]]), .data[[source_var]], .data[[var]]))  # -Y gender-neutral
}
gss_data <- gssrecode(gss_data, vars = c("polhitok", "polabuse", "polmurdr", "polattak", "polescap"), to0 = 2, to1 = 1)
```

### Race and Ethnicity

Childrearing values items use `gss5cats`. The `RACDIF` series uses `-X` variants (extended alternate intro wording on Form 1, informing respondents the questions have been asked for almost 50 years; original intro on Form 2; both modes; negligible gaps). `racdif5` is excluded from the analytic core as it was new in 2024 and not fielded in 2016/2018. Ethnic group comparisons are limited to the White/Black pairings (`wlthwhts`/`wlthblks`, `workwhts`/`workblks`); the Hispanic analogues were not fielded in 2016/2018. Interracial marriage attitude items (`MARBLK`, `MARASIAN`, `MARHISP`, `MARWHT`) use a five-category favor/oppose scale and are recoded with `gss5cats()`.

```{r race-ethnicity}
gss_data <- gss5cats(gss_data, vars = c("helpoth", "obey", "thnkself", "workhard", "popular"))

racdif_vars <- c("racdif1", "racdif2", "racdif3", "racdif4")
for (var in racdif_vars) {
  source_var <- paste0(var, "x")
  gss_data <- gss_data %>%
    mutate(!!sym(var) := if_else(form == 1 & !is.na(.data[[source_var]]), .data[[source_var]], .data[[var]]))  # -X: see section note
}
gss_data <- gssrecode(gss_data, vars = racdif_vars, to0 = 2, to1 = 1)

ethnic_vars <- c("wlthwhts", "wlthblks", "workwhts", "workblks")
gss_data <- gss7cats(gss_data, vars = c(ethnic_vars, "partyid"))

gss_data <- gss3cats(gss_data, vars = "discaff")
gss_data <- gss5cats(gss_data, vars = "wrkwayup")
gss_data <- gss5cats(gss_data, vars = c("marblk", "marasian", "marhisp", "marwht"))
```

### Gender, Sex, and Suicide

`MARSAME` requires a two-step coalesce across a wording discontinuity and a wording experiment. Step 1 bridges `marsame1` (the 2016/2018 cumulative file name) to `marsame` (2022+). The earlier variable used inadvertent wording — "Homosexual couples *have the right* to marry" (missing "should") — making it descriptive rather than normative; post-*Obergefell* (2015), respondents could agree by acknowledging legal reality. The reversion to "should" in `marsame` is perfectly collinear with the FTF-to-web mode transition, so the mode effect estimate for this variable may be confounded by the concurrent wording change and should be interpreted with caution (see Rosenfeld 2023 addendum). Step 2 coalesces `marsamey` (2024, Form 2), a traditional wording experiment replacing "homosexual couple" with "same-sex couple"; the distributional gap is negligible (\~0.03pp FTF, \~0.00pp web).

`DISCAFFW` is excluded from the analytic core: the base variable has zero web-mode observations because all web cases route to the `-V`/`-NV` variants.

```{r gender-sex-suicide}
gss_data <- gss_data |>
  mutate(
    marsame = if_else(is.na(marsame) & !is.na(marsame1), marsame1, marsame),
    marsame = if_else(form == 2 & !is.na(marsamey), marsamey, marsame)
  )

gss_data <- gss5cats(gss_data, vars = "fehire")
gss_data <- gssrecode(gss_data,
  vars = c("fechld", "fepresch", "meovrwrk", "fejobaff", "discaffm", "fefam", "affrmact"),
  to0 = 1:2, to1 = 3:4
)
gss_data <- gss5cats(gss_data, vars = "marsame")  # 5-point agree/disagree scale; gss5cats for consistency

gss_data <- gssrecode(gss_data, vars = c("homosex", "teensex"), to0 = 1:2, to1 = 3:4)
gss_data <- gssrecode(gss_data, vars = "premarsx", to0 = 1:3, to1 = 4)
gss_data <- gssrecode(gss_data, vars = "xmarsex",  to0 = 1,   to1 = 2:4)
gss_data <- gss3cats(gss_data,  vars = "pornlaw")
gss_data <- gssrecode(gss_data, vars = "sexeduc",  to0 = 2, to1 = 1)
gss_data <- gssrecode(gss_data, vars = "pillok",   to0 = 3:4, to1 = 1:2)

# suicide3 not fielded in 2024
suicide_grid_vars <- c("suicide1", "suicide2", "suicide4")
for (var in suicide_grid_vars) {
  source_var <- paste0(var, "g")
  gss_data <- gss_data %>%
    mutate(!!sym(var) := if_else(form == 1 & !is.na(.data[[source_var]]), .data[[source_var]], .data[[var]]))
}
gss_data <- gssrecode(gss_data, vars = c("letdie1", "suicide1", "suicide2", "suicide4"), to0 = 2, to1 = 1)
```

### Miscellaneous Items

Several variables available in later waves are excluded because they were not fielded in 2016/2018: `adoptus`, `immcrime`, `immjobs`, `immfate`, `letinhsp`, and `letinasn`. `letin1a` is retained.

```{r misc-recodes}
misc_yesno_vars <- c("cappun", "gunlaw", "grass", "postlife", "prayer",
                     "richwork", "fear", "reborn", "xmovie")
gss_data <- gssrecode(gss_data, vars = misc_yesno_vars, to0 = 2, to1 = 1)

misc_4cat_vars <- c("spanking", "class", "relpersn", "sprtprsn", "joblose", "health", "satjob")
gss_data <- gssrecode(gss_data, vars = misc_4cat_vars, to0 = 1:2, to1 = 3:4)

misc_5cat_vars <- c("helppoor", "helpnot", "helpsick", "letin1a",
                    "helpblk", "parsol", "kidssol", "goodlife", "finrela")
gss_data <- gss5cats(gss_data, vars = misc_5cat_vars)

misc_depends_vars <- c("fair", "helpful", "trust")
gss_data <- gssrecode(gss_data, vars = misc_depends_vars, to0 = 1, to1 = 2)

gss_data <- gssrecode(gss_data, vars = "eqwlth",   to0 = 1:3, to1 = 4:7)
gss_data <- gssrecode(gss_data, vars = "polviews", to0 = 1:4, to1 = 5:7)
gss_data <- gssrecode(gss_data, vars = "getahead", to0 = 2:3, to1 = 1)
gss_data <- gssrecode(gss_data, vars = "tax",      to0 = 2:3, to1 = 1)
gss_data <- gssrecode(gss_data, vars = "life",     to0 = 1,   to1 = 2:3)

gss_data <- gss3cats(gss_data, vars = c("happy", "satfin", "finalter", "jobfind", "divlaw"))
gss_data <- gssrecode(gss_data, vars = "courts", to0 = c(1, 3), to1 = 2)

gss_data <- gssrecode(gss_data, vars = "attend",   to0 = 0:4, to1 = 5:8)
gss_data <- gssrecode(gss_data, vars = "pray",     to0 = 4:6, to1 = 1:3)
gss_data <- gssrecode(gss_data, vars = "bible",    to0 = 2:4, to1 = 1)
gss_data <- gssrecode(gss_data, vars = "relactiv", to0 = 1:4, to1 = 5:10)
gss_data <- gssrecode(gss_data, vars = "chldidel", to0 = 0:2, to1 = c(3:8))
gss_data <- gssrecode(gss_data, vars = "rank",     to0 = 6:10, to1 = 1:5)
gss_data <- gssrecode(gss_data, vars = "god",      to0 = 1:3,  to1 = 4:6)
```

------------------------------------------------------------------------

## 4. Filter and Export

```{r export}
gss_data_16to24 <- gss_data %>%
  filter(year %in% c(2016, 2018, 2022, 2024))

saveRDS(gss_data_16to24, GSS_RDS_PATH)
```
