---
title: "GSS Variable Recoding and Preparation (2016–2024)"
editor: visual
---

## Overview

Recodes subjective items from the 2016–2024 General Social Survey cumulative file into binary (`c_`-prefixed) outcome variables and saves the result as an RDS for downstream analysis. Three operations are applied per variable:

1.  **Coalescing** — merges split-ballot variants (Form 1/2, visual/non-visual flashcard, `-g`/`-y` suffixes) into a single variable.
2.  **Binarization** — converts ordinal scales into 0/1 outcomes using distribution-based split rules.
3.  **Filtering** — restricts to 2016, 2018, 2022, and 2024.

Only variables in the analytic core (`data/analytic_core.csv`, `analytic_core == 1`) are used in downstream files; recodes for excluded items are omitted here.

------------------------------------------------------------------------

## 1. Setup and Data Loading

```{r setup, message=FALSE, warning=FALSE}
source("00_setup.R")

if (file.exists(GSS_DTA_PATH)) {
  gss_data <- read_dta(GSS_DTA_PATH)
} else {
  stop("Data file not found: ", GSS_DTA_PATH)
}
```

------------------------------------------------------------------------

## 2. Recoding Functions

```{r functions}
#' Replicates Stata's `recode` command for binary variables.
gssrecode <- function(data, vars, to0, to1) {
  for (var in vars) {
    if (!var %in% names(data)) {
      warning(paste0("Variable '", var, "' not found. Skipping."))
      next
    }
    new_var_name <- paste0("c_", var)
    data <- data %>%
      mutate(!!new_var_name := case_when(
        .data[[var]] %in% to0 ~ 0,
        .data[[var]] %in% to1 ~ 1,
        TRUE ~ NA_real_
      ))
  }
  return(data)
}

#' Recodes 3-category variables into a binary split based on distribution.
gss3cats <- function(data, vars) {
  for (var in vars) {
    if (!var %in% names(data)) next
    num1 <- sum(data[[var]] == 1, na.rm = TRUE)
    num3 <- sum(data[[var]] == 3, na.rm = TRUE)
    if (num1 >= num3) {
      data <- gssrecode(data, var, to0 = c(2, 3), to1 = 1)
    } else {
      data <- gssrecode(data, var, to0 = 3, to1 = c(1, 2))
    }
  }
  return(data)
}

#' Recodes 5-category variables into a binary split based on distribution.
gss5cats <- function(data, vars) {
  for (var in vars) {
    if (!var %in% names(data)) next
    num12 <- sum(data[[var]] %in% c(1, 2), na.rm = TRUE)
    num45 <- sum(data[[var]] %in% c(4, 5), na.rm = TRUE)
    if (num12 >= num45) {
      data <- gssrecode(data, var, to0 = c(3, 4, 5), to1 = c(1, 2))
    } else {
      data <- gssrecode(data, var, to0 = c(4, 5), to1 = c(1, 2, 3))
    }
  }
  return(data)
}

#' Recodes 7-category variables into a binary split based on distribution.
gss7cats <- function(data, vars) {
  for (var in vars) {
    if (!var %in% names(data)) next
    num123 <- sum(data[[var]] %in% c(1, 2, 3), na.rm = TRUE)
    num567 <- sum(data[[var]] %in% c(5, 6, 7), na.rm = TRUE)
    if (num123 >= num567) {
      data <- gssrecode(data, var, to0 = c(4, 5, 6, 7), to1 = c(1, 2, 3))
    } else {
      data <- gssrecode(data, var, to0 = c(5, 6, 7), to1 = c(1, 2, 3, 4))
    }
  }
  return(data)
}
```

------------------------------------------------------------------------

## 3. Recoding

### National Spending Items

Coalesces Form 2 variants (`-y` suffix) into main variable names before recoding.

```{r nat-spend}
gss_data <- gss_data %>%
  mutate(
    natspac  = if_else(form == 2 & !is.na(natspacy),  natspacy,  natspac),
    natenvir = if_else(form == 2 & !is.na(natenviy),  natenviy,  natenvir),
    natheal  = if_else(form == 2 & !is.na(nathealy),  nathealy,  natheal),
    natdrug  = if_else(form == 2 & !is.na(natdrugy),  natdrugy,  natdrug),
    nateduc  = if_else(form == 2 & !is.na(nateducy),  nateducy,  nateduc),
    natarms  = if_else(form == 2 & !is.na(natarmsy),  natarmsy,  natarms),
    natrace  = if_else(form == 2 & !is.na(natracey),  natracey,  natrace),
    nataid   = if_else(form == 2 & !is.na(nataidy),   nataidy,   nataid),
    natfare  = if_else(form == 2 & !is.na(natfarey),  natfarey,  natfare),
    natcity  = if_else(form == 2 & !is.na(natcityy),  natcityy,  natcity),
    natcrime = if_else(form == 2 & !is.na(natcrimy),  natcrimy,  natcrime)
  )

natspend_vars <- c("natspac", "natenvir", "natheal", "natcity", "natcrime",
                   "natdrug", "nateduc", "natrace", "natarms", "nataid",
                   "natfare", "natroad", "natsoc", "natchld", "natsci", "natenrgy")
gss_data <- gss3cats(gss_data, vars = natspend_vars)
```

### Confidence and Civil Liberties

Handles Form 1/2 and visual/non-visual (`-v`/`-nv`) flashcard experiments.

```{r civil-liberties}
gss_data <- gss_data %>%
  mutate(
    conlabor = if_else(form == 2 & !is.na(conlabory), conlabory, conlabor),
    colcom   = if_else(form == 2 & !is.na(colcomy),   colcomy,   colcom),
    spkath   = if_else(form == 2 & !is.na(spkathy),   spkathy,   spkath),
    spkrac   = if_else(form == 2 & !is.na(spkracy),   spkracy,   spkrac),
    spkcom   = if_else(form == 2 & !is.na(spkcomy),   spkcomy,   spkcom),
    spkmslm  = if_else(form == 2 & !is.na(spkmslmy),  spkmslmy,  spkmslm),
    libath   = if_else(form == 2 & !is.na(libathy),   libathy,   libath),
    librac   = if_else(form == 2 & !is.na(libracy),   libracy,   librac),
    libcom   = if_else(form == 2 & !is.na(libcomy),   libcomy,   libcom),
    libmslm  = if_else(form == 2 & !is.na(libmslmy),  libmslmy,  libmslm),
    letdie1  = if_else(form == 2 & !is.na(letdie1y),  letdie1y,  letdie1),
    fepol    = if_else(!is.na(fepolv),   fepolv,   if_else(!is.na(fepolnv),   fepolnv,   fepol)),
    racopen  = if_else(!is.na(racopenv), racopenv, if_else(!is.na(racopennv), racopennv, racopen)),
    bible    = if_else(!is.na(biblev),   biblev,   if_else(!is.na(biblenv),   biblenv,   bible)),
    divlaw   = if_else(!is.na(divlawv),  divlawv,  if_else(!is.na(divlawnv),  divlawnv,  divlaw)),
    kidssol  = if_else(!is.na(kidssolv) & kidssolv <= 5, kidssolv,
                       if_else(!is.na(kidssolnv) & kidssolnv <= 5, kidssolnv, kidssol)),
    grass    = if_else(!is.na(grassv),    grassv,    if_else(!is.na(grassnv),    grassnv,    grass)),
    prayer   = if_else(!is.na(prayerv),   prayerv,   if_else(!is.na(prayernv),   prayernv,   prayer)),
    postlife = if_else(!is.na(postlifev), postlifev, if_else(!is.na(postlifenv), postlifenv, postlife)),
    fair     = if_else(!is.na(fairv),     fairv,     if_else(!is.na(fairnv),     fairnv,     fair)),
    helpful  = if_else(!is.na(helpfulv),  helpfulv,  if_else(!is.na(helpfulnv),  helpfulnv,  helpful)),
    trust    = if_else(!is.na(trustv),    trustv,    if_else(!is.na(trustnv),    trustnv,    trust)),
    getahead = if_else(!is.na(getaheadv), getaheadv, if_else(!is.na(getaheadnv), getaheadnv, getahead)),
    courts   = if_else(!is.na(courtsv),   courtsv,   if_else(!is.na(courtsnv),   courtsnv,   courts))
  )

confidence_vars <- c("confinan", "conbus", "conclerg", "coneduc", "confed",
                     "conlabor", "conpress", "conmedic", "contv", "conjudge",
                     "consci", "conlegis", "conarmy")
gss_data <- gss3cats(gss_data, vars = confidence_vars)

gss_data <- gssrecode(gss_data, vars = c("colath", "colrac", "colcom", "colmslm"), to0 = 4, to1 = 5)
gss_data <- gssrecode(gss_data, vars = c("libath", "librac", "libcom", "libmslm", "fepol", "racopen"), to0 = 1, to1 = 2)
gss_data <- gssrecode(gss_data, vars = c("spkath", "spkrac", "spkcom", "spkmslm"), to0 = 1, to1 = 2)
```

### Abortion and Police

Form 1 grid (`-g`) and Form 2 gender-neutral (`-y`) variants.

```{r abortion-police}
abortion_vars <- c("abdefect", "abnomore", "abhlth", "abpoor", "abrape", "absingle", "abany")
for (var in abortion_vars) {
  source_var <- paste0(var, "g")
  gss_data <- gss_data %>%
    mutate(!!sym(var) := if_else(form == 1 & !is.na(.data[[source_var]]), .data[[source_var]], .data[[var]]))
}
gss_data <- gssrecode(gss_data, vars = abortion_vars, to0 = 2, to1 = 1)

police_vars <- c("polhitok", "polabuse", "polattak")
for (var in police_vars) {
  source_var <- paste0(var, "y")
  gss_data <- gss_data %>%
    mutate(!!sym(var) := if_else(form == 2 & !is.na(.data[[source_var]]), .data[[source_var]], .data[[var]]))
}
gss_data <- gssrecode(gss_data, vars = c("polhitok", "polabuse", "polmurdr", "polattak", "polescap"), to0 = 2, to1 = 1)
```

### Race and Ethnicity

```{r race-ethnicity}
# Childrearing
gss_data <- gss5cats(gss_data, vars = c("helpoth", "obey", "thnkself", "workhard", "popular"))

# RACDIF -x variables (Form 1 experimental wording); racdif1-4 only.
# racdif5 (new in 2024, not fielded in 2016/2018) is excluded from analytic core.
racdif_vars <- c("racdif1", "racdif2", "racdif3", "racdif4")
for (var in racdif_vars) {
  source_var <- paste0(var, "x")
  gss_data <- gss_data %>%
    mutate(!!sym(var) := if_else(form == 1 & !is.na(.data[[source_var]]), .data[[source_var]], .data[[var]]))
}
gss_data <- gssrecode(gss_data, vars = racdif_vars, to0 = 2, to1 = 1)

# Ethnic group comparisons (wlthwhts/blks, workwhts/blks only;
# wlthhsps and workhsps not fielded in 2016/2018)
ethnic_vars <- c("wlthwhts", "wlthblks", "workwhts", "workblks")
gss_data <- gss7cats(gss_data, vars = c(ethnic_vars, "partyid"))

gss_data <- gss3cats(gss_data, vars = "discaff")
gss_data <- gss5cats(gss_data, vars = "wrkwayup")
```

### Gender, Sex, and Suicide

```{r gender-sex-suicide}
gss_data <- gss5cats(gss_data, vars = "fehire")
gss_data <- gssrecode(gss_data,
  vars = c("fechld", "fepresch", "meovrwrk", "fejobaff", "discaffm", "discaffw", "fefam", "affrmact"),
  to0 = 1:2, to1 = 3:4
)

gss_data <- gssrecode(gss_data, vars = c("homosex", "teensex"), to0 = 1:2, to1 = 3:4)
gss_data <- gssrecode(gss_data, vars = "premarsx", to0 = 1:3, to1 = 4)
gss_data <- gssrecode(gss_data, vars = "xmarsex",  to0 = 1,   to1 = 2:4)
gss_data <- gss3cats(gss_data,  vars = "pornlaw")
gss_data <- gssrecode(gss_data, vars = "sexeduc",  to0 = 2, to1 = 1)
gss_data <- gssrecode(gss_data, vars = "pillok",   to0 = 3:4, to1 = 1:2)

# Suicide grid variables (suicide3 not fielded in 2024)
suicide_grid_vars <- c("suicide1", "suicide2", "suicide4")
for (var in suicide_grid_vars) {
  source_var <- paste0(var, "g")
  gss_data <- gss_data %>%
    mutate(!!sym(var) := if_else(form == 1 & !is.na(.data[[source_var]]), .data[[source_var]], .data[[var]]))
}
gss_data <- gssrecode(gss_data, vars = c("letdie1", "suicide1", "suicide2", "suicide4"), to0 = 2, to1 = 1)
```

### Miscellaneous Items

```{r misc-recodes}
gss_data <- gss_data %>%
  mutate(xmovie = if_else(form == 2 & !is.na(xmoviey), xmoviey, xmovie))

misc_yesno_vars <- c("cappun", "gunlaw", "grass", "postlife", "prayer",
                     "richwork", "fear", "reborn", "xmovie")
gss_data <- gssrecode(gss_data, vars = misc_yesno_vars, to0 = 2, to1 = 1)

misc_4cat_vars <- c("spanking", "class", "relpersn", "sprtprsn", "joblose", "health", "satjob")
gss_data <- gssrecode(gss_data, vars = misc_4cat_vars, to0 = 1:2, to1 = 3:4)

# letin1a retained; letinhsp and letinasn excluded (not fielded in 2016/2018)
misc_5cat_vars <- c("helppoor", "helpnot", "helpsick", "letin1a",
                    "helpblk", "parsol", "kidssol", "goodlife", "finrela")
gss_data <- gss5cats(gss_data, vars = misc_5cat_vars)

misc_depends_vars <- c("fair", "helpful", "trust")
gss_data <- gssrecode(gss_data, vars = misc_depends_vars, to0 = 1, to1 = 2)

gss_data <- gssrecode(gss_data, vars = "eqwlth",   to0 = 1:3, to1 = 4:7)
gss_data <- gssrecode(gss_data, vars = "polviews", to0 = 1:4, to1 = 5:7)
gss_data <- gssrecode(gss_data, vars = "getahead", to0 = 2:3, to1 = 1)
gss_data <- gssrecode(gss_data, vars = "tax",      to0 = 2:3, to1 = 1)
gss_data <- gssrecode(gss_data, vars = "life",     to0 = 1,   to1 = 2:3)

# adoptus, immcrime, immjobs, immfate excluded (not fielded in 2016/2018)
# hapmar, hapcohab excluded (conditionally fielded — structural skip confounded with mode)
gss_data <- gss3cats(gss_data, vars = c("happy", "satfin", "finalter", "jobfind", "divlaw"))
gss_data <- gssrecode(gss_data, vars = "courts", to0 = c(1, 3), to1 = 2)

gss_data <- gssrecode(gss_data, vars = "attend",   to0 = 0:4, to1 = 5:8)
gss_data <- gssrecode(gss_data, vars = "pray",     to0 = 4:6, to1 = 1:3)
gss_data <- gssrecode(gss_data, vars = "bible",    to0 = 2:4, to1 = 1)
gss_data <- gssrecode(gss_data, vars = "relactiv", to0 = 1:4, to1 = 5:10)
gss_data <- gssrecode(gss_data, vars = "chldidel", to0 = 0:2, to1 = c(3:8))
gss_data <- gssrecode(gss_data, vars = "rank",     to0 = 6:10, to1 = 1:5)
gss_data <- gssrecode(gss_data, vars = "god",      to0 = 1:3,  to1 = 4:6)
```

------------------------------------------------------------------------

## 4. Filter and Export

```{r export}
gss_data_16to24 <- gss_data %>%
  filter(year %in% c(2016, 2018, 2022, 2024))

saveRDS(gss_data_16to24, GSS_RDS_PATH)
```
