---
title: "GSS Mode Reweighting (2022–2024)"
editor: visual
---

## Overview

Computes mode-specific post-stratification weights for the 2022 and 2024 GSS using iterative proportional fitting (raking):

-   **FTF/Phone weight** (`wtssps_f2fp`): Rakes the FTF/Phone subsample to full-sample population margins.
-   **Web weight** (`wtssps_web`): Rakes the Web subsample to full-sample population margins.

Hot-deck imputation handles missing values before raking to prevent case loss. Output is written to `data/gss20222024_modeweights.dta`.

------------------------------------------------------------------------

## Setup

### Load libraries used

```{r setup, message=FALSE, warning=FALSE}
source("00_setup.R")

library(gt)

out_tables <- "outputs/tables"
dir.create(out_tables, recursive = TRUE, showWarnings = FALSE)
```

### Load Data

We load the data here **once** to serve both the diagnostics and the reweighting steps. The RDS file is generated by `01_data_preparation.qmd` and contains all raw variables plus binarized outcome variables.

```{r load-data}

# Load GSS data from RDS (output of 01_data_preparation.qmd)
if (!file.exists(GSS_RDS_PATH)) stop("RDS file not found: ", GSS_RDS_PATH)
gss_full <- readRDS(GSS_RDS_PATH)

# Create year-specific subsets for reweighting
df2022 <- gss_full %>% filter(year == 2022)
df2024 <- gss_full %>% filter(year == 2024)

# Create a combined dataset specifically for diagnostics
gss_combined_diag <- bind_rows(df2022, df2024)
```

## Reweighting function

```{r define-functions}
## REWEIGHTING FUNCTION ----
# Rakes a mode subsample of the GSS to the full-sample (wtssps-weighted) population
# margins. Takes two arguments:
#   `df`               — full year-specific GSS data frame (e.g., df2022)
#   `subset_parameter` — string expression selecting the mode subsample (e.g., "mode %in% c(1, 2)")
# Returns the subsetted data frame with two added weight columns:
#   `raked_weight`    — raw raking weight
#   `rescaled_weight` — raked_weight rescaled so the mean equals 1

reweight <- function(df, subset_parameter) {
  # 7 variables used for raking
  # REVISED (HISP/RACE): Replaced w_hispanic and w_race with w_hisp_race
  vars <- c("w_married", "w_born", "w_male", "w_degree", "w_agegroup", "w_region", "w_hisp_race")

  # clean up data and create the 8 variables for weighting
  df_clean <- df %>%
    filter(
      sample != 4,
      sample != 5,
      sample != 7,
      !is.na(wtssps)  # excludes the supplemental oversample collected using NORC's AmeriSpeak platform
    ) %>%
    mutate(
      w_married = case_when(
        marital == 1 ~ 1,
        marital %in% c(2:5) ~ 0,
      ),
      w_born = case_when(
        born == 1 ~ 1,
        born == 2 ~ 0
      ),
      w_male = case_when(
        sex == 1 ~ 1,
        sex == 2 ~ 0
      ),
      # REVISED (HISP/RACE): Removed w_race
      # REVISED (HISP/RACE): Added w_hisp_race to match the 8 categories from the 2024 codebook
      # This uses 'hispanic' (1=No, 2-5=Yes) and 'racecen1'/'racecen2'
      w_hisp_race = case_when(
        hispanic %in% c(2:5) ~ 1, # Category 1: Hispanic
        hispanic == 1 & racecen1 == 1 & is.na(racecen2) ~ 2, # Category 2: Non-Hispanic, White Alone
        hispanic == 1 & racecen1 == 2 & is.na(racecen2) ~ 3, # Category 3: Non-Hispanic, Black Alone
        hispanic == 1 & racecen1 == 3 & is.na(racecen2) ~ 4, # Category 4: Non-Hispanic, AI/AN Alone
        hispanic == 1 & racecen1 == 4 & is.na(racecen2) ~ 5, # Category 5: Non-Hispanic, Asian Alone
        hispanic == 1 & racecen1 == 5 & is.na(racecen2) ~ 6, # Category 6: Non-Hispanic, NH/PI Alone
        hispanic == 1 & racecen1 == 6 & is.na(racecen2) ~ 7, # Category 7: Non-Hispanic, Other Single Race
        hispanic == 1 & !is.na(racecen2) ~ 8,              # Category 8: Non-Hispanic, Multiple Races
      ),
      w_degree = case_when(
        degree == 0 ~ 1,
        degree %in% c(1:2) ~ 2,
        degree %in% c(3:4) ~ 3,
      ),
      # REVISED (AGEGROUP): Matched bins to codebook excerpt
      w_agegroup = case_when(
        age %in% c(18:29) ~ 1,
        age %in% c(30:39) ~ 2,
        age %in% c(40:49) ~ 3,
        age %in% c(50:64) ~ 4,
        age %in% c(65:89) ~ 5, # Corrected max age to 89 per codebook
      ),
      w_region = region
    )


  ## HOT DECK IMPUTATION
  # there are some missingness across the sample for the eight variables used for weighting,
  # which leads to these observations from being dropped in the re-weighting process,
  # so hot deck imputation is used to impute the missing values
  # REVISED (HISP/RACE): The 'vars' vector is updated, so this now imputes w_hisp_race
  df_clean <- hotdeck(df_clean, vars)

  ## PROPORTIONS AFTER WEIGHTING
  # full sample with original post-stratification GSS weights (WTSSPS)
  full_design <- svydesign(ids = ~1, data = df_clean, weights = ~wtssps)

  # proportions from the full sample using original post-stratification GSS weights (WTSSPS)
  full_props <- list(
    w_married = svytable(~ w_married, full_design),
    w_born = svytable(~ w_born, full_design),
    w_male = svytable(~ w_male, full_design),
    w_degree = svytable(~ w_degree, full_design),
    w_agegroup = svytable(~ w_agegroup, full_design),
    w_region = svytable(~ w_region, full_design),
    w_hisp_race = svytable(~ w_hisp_race, full_design)
  )


  # subset sample (ie f2f/phone sample) with original post-stratification GSS weights (WTSSPS)
  df_subset <- subset(df_clean, eval(parse(text = subset_parameter)))
  subset_design <- svydesign(ids = ~1, data = df_subset, weights = ~wtssps)



  ## REWEIGHTING
  # reweighting subset sample using the rake() function
  # based on the proportions of the full sample
  raked_design <- rake(
    design = subset_design,
    # REVISED (HISP/RACE): Updated list of margins
    sample.margins = list(~w_married, ~w_born, ~w_male, ~w_degree, ~w_agegroup, ~w_region, ~w_hisp_race),
    population.margins = full_props # using the full sample proportions
  )

  ## CREATE NEW DATAFRAME
  # new df with raked_weight and rescaled_weight (which is rescaled so the mean is 1)
  df_new <- raked_design$variables %>%
    bind_cols(as.data.frame(weights(raked_design))) %>%
    rename(raked_weight = `weights(raked_design)`) %>% # new weights
    mutate(rescaled_weight = raked_weight / mean(raked_weight)) # rescaled weights


  return(df_new)
}
```

## DIAGNOSTICS: Demographic Balance Check

This section uses the data loaded in Setup to visualize selection bias. We compare the **Target** (Full Sample, Weighted) against the **F2F** and **Web** subsamples, looking at both their **Unweighted** (raw) and **Original Weighted** (`wtssps`) distributions.

```{r diagnostics, message=FALSE, warning=FALSE, fig.height=10, fig.width=10}
# --- 1. PREP DIAGNOSTIC DATA ---
# This reuses the 'gss_combined_diag' object from the Setup chunk.
# We create labeled versions of the variables for easier plotting.

prep_diagnostic_data <- function(df) {
  # Clean and label
  df_labeled <- df %>%
    filter(sample != 4, sample != 5, sample != 7, !is.na(wtssps)) %>%
    mutate(
      w_married = case_when(marital == 1 ~ "Married", marital %in% 2:5 ~ "Not Married"),
      w_born = case_when(born == 1 ~ "Born in US", born == 2 ~ "Born Outside the US"),
      w_male = case_when(sex == 1 ~ "Male", sex == 2 ~ "Female"),
      w_degree = case_when(degree == 0 ~ "Less than High School", degree %in% 1:2 ~ "High School", degree %in% 3:4 ~ "Bachelor's or More"),
      w_agegroup = case_when(
        age %in% 18:29 ~ "18-29", age %in% 30:39 ~ "30-39", age %in% 40:49 ~ "40-49",
        age %in% 50:64 ~ "50-64", age %in% 65:89 ~ "65+"
      ),
      w_hisp_race = case_when(
        hispanic %in% c(2:5)                                    ~ "Hispanic",
        hispanic == 1 & racecen1 == 1 & is.na(racecen2)        ~ "Non-Hispanic White Alone",
        hispanic == 1 & racecen1 == 2 & is.na(racecen2)        ~ "Non-Hispanic Black Alone",
        hispanic == 1 & racecen1 == 3 & is.na(racecen2)        ~ "Non-Hispanic AIAN Alone",
        hispanic == 1 & racecen1 %in% c(4:10) & is.na(racecen2) ~ "Non-Hispanic Asian Alone",
        hispanic == 1 & racecen1 == 14 & is.na(racecen2)       ~ "Non-Hispanic NHPI Alone",
        hispanic == 1 & racecen1 %in% c(15, 16) & is.na(racecen2) ~ "Non-Hispanic Other Race Alone",
        hispanic == 1 & !is.na(racecen2)                        ~ "Non-Hispanic Multiple Races"
      ),
      w_region = as.character(region)
    ) %>%
    drop_na(w_married, w_born, w_male, w_degree, w_agegroup, w_hisp_race, w_region)

  return(df_labeled)
}

diag_data <- prep_diagnostic_data(gss_combined_diag)

# --- 2. CALCULATE PROPORTIONS ---

calc_diag_props <- function(data, group_label, weight_col = NULL) {
  vars <- c("w_married", "w_born", "w_male", "w_degree", "w_agegroup", "w_hisp_race")

  if (is.null(weight_col)) {
    des <- svydesign(ids = ~1, data = data)
  } else {
    des <- svydesign(ids = ~1, data = data, weights = data[[weight_col]])
  }

  lapply(vars, function(v) {
    as.data.frame(prop.table(svytable(as.formula(paste0("~", v)), des))) %>%
      rename(value = 1, prop = Freq) %>%
      mutate(variable = v, group = group_label)
  }) %>% bind_rows()
}

# A. Target: Full Sample (wtssps)
props_target <- calc_diag_props(diag_data, "1. Target (Full, wtssps)", "wtssps")

# B. F2F: Unweighted
props_f2f_unw <- calc_diag_props(diag_data %>% filter(mode %in% 1:2), "2. F2F (Unweighted)", NULL)

# C. Web: Unweighted
props_web_unw <- calc_diag_props(diag_data %>% filter(mode == 4), "3. Web (Unweighted)", NULL)

# D. F2F: Original Weight (wtssps)
props_f2f_wt <- calc_diag_props(diag_data %>% filter(mode %in% 1:2), "4. F2F (wtssps)", "wtssps")

# E. Web: Original Weight (wtssps)
props_web_wt <- calc_diag_props(diag_data %>% filter(mode == 4), "5. Web (wtssps)", "wtssps")

# Combine
all_diag_props <- bind_rows(props_target, props_f2f_unw, props_web_unw, props_f2f_wt, props_web_wt)

# --- 3. OUTPUT TABLE ---
all_diag_props %>%
  pivot_wider(names_from = group, values_from = prop) %>%
  mutate(across(where(is.numeric), \(x) round(x, 3))) %>%
  gt() %>%
  tab_header(title = "Demographic Proportions: Target vs. Mode Subsamples")

# --- 4. VISUALIZATION ---
ggplot(all_diag_props, aes(x = value, y = prop, fill = group)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  facet_wrap(~variable, scales = "free", ncol = 2) +
  coord_flip() +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal() +
  labs(
    title = "Demographic Balance by Mode and Weighting Scheme",
    y = "Proportion", x = "", fill = "Group"
  ) +
  theme(legend.position = "bottom", strip.text = element_text(face = "bold"))
```

## Compute weights

Note: We use the `df2022` and `df2024` objects created in the Setup chunk.

### Reweight phone only cases

```{r compute-f2f}
SUBSET_PARAMETER <- "mode %in% c(1, 2)"  # this is for f2f/phone sample of GSS
reweight_2022_f2fp <- reweight(df2022, SUBSET_PARAMETER) |>
  rename(wtssps_f2fp = rescaled_weight) |>
  select(id, year, wtssps_f2fp)
reweight_2024_f2fp <- reweight(df2024, SUBSET_PARAMETER) |>
  rename(wtssps_f2fp = rescaled_weight) |>
  select(id, year, wtssps_f2fp)
```

### Reweight web only cases

```{r compute-web}
SUBSET_PARAMETER <- "mode %in% c(4)"  # this is for web sample of GSS
reweight_2022_web <- reweight(df2022, SUBSET_PARAMETER) |>
  rename(wtssps_web = rescaled_weight) |>
  select(id, year, wtssps_web)
reweight_2024_web <- reweight(df2024, SUBSET_PARAMETER) |>
  rename(wtssps_web = rescaled_weight) |>
  select(id, year, wtssps_web)
```

### Combine into one dataset

```{r save-output}
combined <- bind_rows(reweight_2022_f2fp, reweight_2022_web, reweight_2024_f2fp, reweight_2024_web)
write_dta(combined, WEIGHTS_PATH)
```
