---
title: "GSS Sample Composition by Mode (2016–2024)"
editor: visual
---

## Overview

This file produces two paper figures documenting mode-related composition differences in the 2022–2024 GSS multimode waves:

-   Stacked bar chart of GSS sample sizes by mode and year (2016, 2018, 2022, 2024).
-   Dot plot of unweighted demographic composition by mode (2022–2024) against population benchmarks derived from `wtssps` on the full sample.

In-text statistics cited in §2.1 are reproducible from the `unweighted_props` object printed in the verification chunk below.

------------------------------------------------------------------------

## Setup

```{r setup, message=FALSE, warning=FALSE}
source("00_setup.R")

# File-specific libraries not covered by 00_setup.R
library(gt)
library(scales)

# Output directory
out_fig <- "outputs/figures"
dir.create(out_fig, recursive = TRUE, showWarnings = FALSE)
```

------------------------------------------------------------------------

## Load Data

```{r load-data, message=FALSE, warning=FALSE}
if (!file.exists(GSS_RDS_PATH)) stop("RDS file not found: ", GSS_RDS_PATH)
gss_full <- readRDS(GSS_RDS_PATH)
```

------------------------------------------------------------------------

## Define Variables

```{r define-vars, message=FALSE, warning=FALSE}
# Filter to analysis years, apply canonical demographic recoding, and construct mode variables.

gss_analysis <- gss_full %>%
  filter(year %in% c(2016, 2018, 2022, 2024)) %>%
  recode_demographics() %>%
  mutate(
    # Survey mode — three-level factor (unique to 03_composition for stacked bar chart)
    mode_label = case_when(
      year %in% c(2016, 2018) & mode == 1 ~ "FTF",
      year %in% c(2016, 2018) & mode == 2 ~ "Phone",
      year %in% c(2022, 2024) & mode == 1 ~ "FTF",
      year %in% c(2022, 2024) & mode == 2 ~ "Phone",
      year %in% c(2022, 2024) & mode == 4 ~ "Web",
      TRUE                               ~ NA_character_
    ),

    # Binary mode indicator (2022–2024 only) — derived from canonical mode_fctr
    mode_bin = case_when(
      mode %in% c(1, 2) ~ "FTF/Phone",
      mode == 4         ~ "Web",
      TRUE              ~ NA_character_
    )
  )
```

------------------------------------------------------------------------

## Sample Sizes by Mode and Year

```{r sample-sizes-chart, message=FALSE, warning=FALSE, fig.width=8, fig.height=5}
# Counts by year and mode_label
fig1_data <- gss_analysis %>%
  filter(!is.na(mode_label)) %>%
  count(year, mode_label) %>%
  mutate(
    year       = factor(year),
    mode_label = factor(mode_label, levels = c("Web", "Phone", "FTF"))
  )

fig1 <- ggplot(fig1_data, aes(x = year, y = n, fill = mode_label)) +
  geom_col(width = 0.65) +
  geom_text(
    aes(label = scales::comma(n)),
    position = position_stack(vjust = 0.5),
    colour   = "white",
    fontface = "bold",
    size     = 5
  ) +
  scale_fill_manual(
  values = c("FTF" = "#D95F02", "Phone" = "#F4A261", "Web" = "#1B9E77"),
  name   = "Mode"
) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.05))) +
  labs(
    title    = "GSS Sample Sizes by Mode and Year",
    subtitle = "Unweighted respondent counts; 2016 and 2018 are single-mode (FTF/Phone)",
    x        = "Survey Year",
    y        = "Number of Respondents"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position   = "bottom",
    panel.grid.major.x = element_blank(),
    plot.title        = element_text(face = "bold")
  )

fig1

ggsave(
  filename = file.path(out_fig, "sample_sizes.png"),
  plot     = fig1,
  width    = 8, height = 5, dpi = 300
)
```

------------------------------------------------------------------------

## Unweighted Demographic Composition by Mode

```{r demographic-composition-dotplot, message=FALSE, warning=FALSE, fig.width=10, fig.height=10}
# ── A. Population benchmarks ──────────────────────────────────────────────────
# Derived from wtssps on the full 2022–2024 sample (both modes combined).
# This mirrors the NORC post-stratification target and serves as the
# reference population distribution.

gss_2224 <- gss_analysis %>%
  filter(year %in% c(2022, 2024), !is.na(wtssps))

compute_wtd_props <- function(data, var, weight_col = "wtssps") {
  data %>%
    filter(!is.na(.data[[var]])) %>%
    group_by(value = .data[[var]]) %>%
    summarise(wt_sum = sum(.data[[weight_col]], na.rm = TRUE), .groups = "drop") %>%
    mutate(
      prop     = wt_sum / sum(wt_sum),
      variable = var
    )
}

demo_vars <- c("marital_cat", "f_born", "f_sex", "degree_cat", "agegrp", "w_hisp_race")

pop_benchmarks <- map_dfr(demo_vars, ~ compute_wtd_props(gss_2224, .x)) %>%
  mutate(group = "Population")

# ── B. Unweighted mode-specific proportions ───────────────────────────────────
gss_2224_modes <- gss_2224 %>%
  filter(!is.na(mode_bin))

compute_unwtd_props <- function(data, var, group_label) {
  data %>%
    filter(!is.na(.data[[var]])) %>%
    count(value = .data[[var]]) %>%
    mutate(
      prop     = n / sum(n),
      variable = var,
      group    = group_label
    ) %>%
    select(-n)
}

unweighted_props <- bind_rows(
  map_dfr(demo_vars, ~ compute_unwtd_props(
    filter(gss_2224_modes, mode_bin == "FTF/Phone"), .x, "FTF/Phone"
  )),
  map_dfr(demo_vars, ~ compute_unwtd_props(
    filter(gss_2224_modes, mode_bin == "Web"), .x, "Web"
  ))
)

# ── C. Assemble plot data ─────────────────────────────────────────────────────
plot_data_unweighted <- bind_rows(
  unweighted_props,
  pop_benchmarks %>% select(variable, value, prop, group)
) %>%
  mutate(
    group = factor(group, levels = c("FTF/Phone", "Web", "Population")),
    facet_label = case_when(
      variable == "marital_cat"  ~ "Marital Status",
      variable == "f_born"       ~ "Nativity",
      variable == "f_sex"        ~ "Sex",
      variable == "degree_cat"   ~ "Education",
      variable == "agegrp"       ~ "Age",
      variable == "w_hisp_race"  ~ "Race/Ethnicity",
      TRUE                       ~ variable
    )
  )

# ── D. Plot ───────────────────────────────────────────────────────────────────
fig2 <- ggplot(
  plot_data_unweighted,
  aes(x = prop, y = reorder(value, prop), colour = group, shape = group)
) +
  geom_point(size = 4, alpha = 0.85) +
  facet_wrap(~ facet_label, scales = "free_y", ncol = 2) +
  scale_colour_manual(
    values = c("FTF/Phone" = "#D95F02", "Web" = "#1B9E77", "Population" = "#000000"),
    name   = NULL
  ) +
  scale_shape_manual(
    values = c("FTF/Phone" = 16, "Web" = 17, "Population" = 18),
    name   = NULL
  ) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1),
                     expand = expansion(mult = c(0.05, 0.1))) +
  labs(
    title    = "Unweighted Demographic Composition by Mode (2022–2024)",
    subtitle = "FTF/Phone (circle), Web (triangle), Population benchmark from wtssps (diamond)",
    x        = "Proportion",
    y        = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position  = "bottom",
    strip.text       = element_text(face = "bold"),
    plot.title       = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

fig2

ggsave(
  filename = file.path(out_fig, "demographic_composition.png"),
  plot     = fig2,
  width    = 10, height = 10, dpi = 300
)
```

------------------------------------------------------------------------

## Age Means by Mode

```{r age-means-by-mode, message=FALSE, warning=FALSE}
age_means <- gss_2224_modes %>%
  filter(!is.na(age)) %>%
  group_by(mode_bin) %>%
  summarise(
    n = n(),
    mean_age = mean(age, na.rm = TRUE),
    sd_age = sd(age, na.rm = TRUE),
    min_age = min(age, na.rm = TRUE),
    max_age = max(age, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    mode_bin = factor(mode_bin, levels = c("FTF/Phone", "Web"))
  ) %>%
  arrange(mode_bin)

age_means %>%
  gt() %>%
  tab_header(
    title    = "Age Summary Statistics by Mode (2022–2024)",
    subtitle = "Unweighted respondent data"
  ) %>%
  fmt_number(columns = c(mean_age, sd_age), decimals = 1) %>%
  fmt_integer(columns = c(n, min_age, max_age)) %>%
  cols_label(
    mode_bin  = "Mode",
    n         = "N",
    mean_age  = "Mean Age",
    sd_age    = "Std Dev",
    min_age   = "Min",
    max_age   = "Max"
  )

# Export age means for §2.1 verification
write.csv(
  age_means,
  file.path(out_fig, "../data/age_means_by_mode.csv"),
  row.names = FALSE
)
```

------------------------------------------------------------------------

## Population Benchmark Values by Year

```{r pop-benchmarks-by-year-table, message=FALSE, warning=FALSE}
# Compute weighted proportions by year
pop_benchmarks_by_year <- gss_analysis %>%
  filter(year %in% c(2016, 2018, 2022, 2024), !is.na(wtssps)) %>%
  pivot_longer(
    cols = all_of(demo_vars),
    names_to = "variable",
    values_to = "value"
  ) %>%
  filter(!is.na(value)) %>%
  group_by(year, variable, value) %>%
  summarise(
    wt_sum = sum(wtssps, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(year, variable) %>%
  mutate(prop = wt_sum / sum(wt_sum)) %>%
  ungroup() %>%
  mutate(
    variable_label = case_when(
      variable == "marital_cat"  ~ "Marital Status",
      variable == "f_born"       ~ "Nativity",
      variable == "f_sex"        ~ "Sex",
      variable == "degree_cat"   ~ "Education",
      variable == "agegrp"       ~ "Age",
      variable == "w_hisp_race"  ~ "Race/Ethnicity",
      TRUE                       ~ variable
    )
  ) %>%
  select(variable_label, value, year, prop) %>%
  pivot_wider(
    names_from = year,
    values_from = prop,
    names_prefix = "Year_"
  ) %>%
  rename(Category = value)

pop_benchmarks_by_year %>%
  gt(groupname_col = "variable_label") %>%
  tab_header(
    title    = "Population Benchmark Values by Year",
    subtitle = "Weighted proportions using wtssps"
  ) %>%
  fmt_percent(columns = c(Year_2016, Year_2018, Year_2022, Year_2024), decimals = 1) %>%
  cols_label(
    Category = "Category",
    Year_2016 = "2016",
    Year_2018 = "2018",
    Year_2022 = "2022",
    Year_2024 = "2024"
  )

# Export population benchmarks for §2.1 verification
write.csv(
  pop_benchmarks_by_year,
  file.path(out_fig, "../data/population_benchmarks_by_year.csv"),
  row.names = FALSE
)
```

------------------------------------------------------------------------

## Verification: In-Text Statistics (§2.1)

The chunk below explicitly prints the values cited in the paper for easy verification during revision.

```{r verify-intext-stats, message=FALSE, warning=FALSE}
verify_stats <- unweighted_props %>%
  filter(
    (variable == "degree_cat"  & value == "Bachelor's or More") |
    (variable == "marital_cat" & value == "Married")            |
    (variable == "w_hisp_race" & value == "Non-Hispanic White Alone")
  ) %>%
  select(group, variable, value, prop) %>%
  mutate(prop_pct = round(prop * 100, 1)) %>%
  arrange(variable, group)

verify_stats %>%
  gt() %>%
  tab_header(
    title    = "In-Text Statistics Verification (§2.1)",
    subtitle = "Unweighted proportions by mode, 2022–2024"
  ) %>%
  fmt_percent(columns = prop, decimals = 1) %>%
  cols_label(
    group    = "Mode",
    variable = "Variable",
    value    = "Category",
    prop     = "Proportion",
    prop_pct = "Pct (%)"
  )
```
