---
title: "GSS Mode Gaps: Selection and Substantive Effects (2022–2024)"
editor: visual
---

## Overview

This file produces all §2 outputs for the paper:

-   **§2.1 — Section 1**: Logistic regression predicting web vs. FTF/Phone completion, with odds ratios and a Shapley value decomposition of predictor importance.
-   **§2.2 — Section 2**: Mode gaps in substantive survey responses before and after demographic adjustment across 137 outcome variables.

------------------------------------------------------------------------

## Setup

```{r setup, message=FALSE, warning=FALSE}
source("00_setup.R")

# File-specific libraries not covered by 00_setup.R
library(broom)
library(gt)
library(dominanceanalysis)

# Output directories
out_fig    <- "outputs/figures"
out_tables <- "outputs/tables"
out_data   <- "outputs/data"
for (d in c(out_fig, out_tables, out_data)) {
  dir.create(d, recursive = TRUE, showWarnings = FALSE)
}
```

------------------------------------------------------------------------

## Shared Data Loading and Recodes

```{r load-data, message=FALSE, warning=FALSE}
if (!file.exists(GSS_RDS_PATH)) stop("RDS not found: ", GSS_RDS_PATH)
gss_data <- readRDS(GSS_RDS_PATH)
```

```{r shared-recodes, message=FALSE, warning=FALSE}
# Shared analytic sample for both sections.
# recode_demographics() from 00_setup.R supplies: mode_fctr, agegrp,
# degree_cat, marital_cat, w_hisp_race, f_born, f_sex, f_region.
analysis_data <- gss_data %>%
  filter(year %in% c(2022, 2024)) %>%
  recode_demographics() %>%
  filter(
    !is.na(wtssps),
    !is.na(mode_fctr),
    !is.na(year),
    !is.na(agegrp),
    !is.na(degree_cat),
    !is.na(marital_cat),
    !is.na(w_hisp_race),
    !is.na(f_born),
    !is.na(f_sex),
    !is.na(f_region)
  )

tibble(
  Statistic = c("Total (complete cases)", "FTF/Phone", "Web"),
  N = c(
    nrow(analysis_data),
    sum(analysis_data$mode_fctr == "f2fp", na.rm = TRUE),
    sum(analysis_data$mode_fctr == "web",  na.rm = TRUE)
  )
) |> gt() |> tab_header(title = "Analytic sample size")
```

------------------------------------------------------------------------

## Section 1: Predicting Mode of Completion (§2.1)

This section examines whether and how strongly demographic characteristics predict a respondent's likelihood of completing via Web versus FTF/Phone. The full logistic model estimates ORs for each predictor; the Shapley decomposition (Table 1) quantifies each predictor's unique contribution to overall model fit.

### Logistic Regression and Shapley Decomposition Predicting Web Completion

```{r full-logistic-model, message=FALSE, warning=FALSE}
# Section 1 uses a numeric 0/1 outcome (mode_fctr in recode_demographics()
# is a factor; recode here for glm compatibility in the selection model).
weighting_vars <- c("marital_cat", "agegrp", "degree_cat",
                    "w_hisp_race", "f_born", "f_sex", "f_region")

selection_data <- analysis_data %>%
  mutate(
    mode_bin = if_else(mode_fctr == "web", 1L, 0L)
  )

full_model_formula <- as.formula(
  paste("mode_bin ~ year +", paste(weighting_vars, collapse = " + "))
)

full_model <- glm(
  full_model_formula,
  family = binomial(link = "logit"),
  data   = selection_data
)

tidy(full_model, conf.int = TRUE) %>%
  gt() %>%
  tab_header(
    title    = "Logistic Regression: Predictors of Web Completion",
    subtitle = "Log-odds coefficients (mode_bin ~ year + demographics), GSS 2022–2024"
  ) %>%
  fmt_number(columns = c(estimate, std.error, statistic, conf.low, conf.high), decimals = 3) %>%
  fmt_number(columns = p.value, decimals = 4) %>%
  cols_label(
    term      = "Term",
    estimate  = "Log-Odds",
    std.error = "SE",
    statistic = "z",
    p.value   = "p",
    conf.low  = "CI Low",
    conf.high = "CI High"
  )
```

```{r dominance-analysis, message=FALSE, warning=FALSE}
# PAPER: Table 1 — Shapley decomposition
da_results <- dominanceAnalysis(full_model, target = "r2.m")
da_stats   <- da_results$contribution.average$r2.m

da_df <- data.frame(
  Variable     = names(da_stats),
  Contribution = as.numeric(da_stats)
) %>%
  mutate(
    Rank           = rank(-Contribution),
    Percent_of_Fit = (Contribution / sum(Contribution)) * 100
  ) %>%
  arrange(desc(Contribution))

da_df %>%
  gt() %>%
  tab_header(
    title    = "Shapley Decomposition of Mode Selection Model",
    subtitle = "Average marginal R² contribution per predictor (McFadden)"
  ) %>%
  fmt_number(columns = Contribution, decimals = 4) %>%
  fmt_number(columns = Percent_of_Fit, decimals = 1) %>%
  fmt_integer(columns = Rank) %>%
  cols_label(
    Variable       = "Predictor",
    Contribution   = "Avg. Marginal R²",
    Rank           = "Rank",
    Percent_of_Fit = "% of Fit"
  )
```

```{r table1-combined, message=FALSE, warning=FALSE}
# PAPER: Table 1 — Merge ORs and Shapley values

# 1. Extract full-model ORs collapsed to variable level
or_df <- tidy(full_model, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    variable = case_when(
      str_starts(term, "degree_cat")  ~ "degree_cat",
      str_starts(term, "marital_cat") ~ "marital_cat",
      str_starts(term, "agegrp")      ~ "agegrp",
      str_starts(term, "year")        ~ "year",
      str_starts(term, "w_hisp_race") ~ "w_hisp_race",
      str_starts(term, "f_born")      ~ "f_born",
      str_starts(term, "f_region")    ~ "f_region",
      str_starts(term, "f_sex")       ~ "f_sex",
      TRUE ~ term
    )
  )

or_summary <- or_df %>%
  group_by(variable) %>%
  summarise(
    OR      = exp(mean(log(estimate))),
    CI_low  = min(conf.low),
    CI_high = max(conf.high),
    .groups = "drop"
  ) %>%
  mutate(
    `95% CI` = paste0(
      "[", formatC(CI_low,  digits = 2, format = "f"), ", ",
           formatC(CI_high, digits = 2, format = "f"), "]"
    )
  )

# 2. Join with Shapley values
table1_data <- or_summary %>%
  left_join(
    da_df %>% select(Variable, Percent_of_Fit, Rank) %>%
      rename(variable = Variable),
    by = "variable"
  ) %>%
  arrange(desc(Percent_of_Fit))

# 3. Render as gt table
gt_table1 <- table1_data %>%
  select(variable, OR, `95% CI`, Percent_of_Fit, Rank) %>%
  gt() %>%
  tab_header(
    title    = "Predictors of Web Survey Completion",
    subtitle = "Logistic regression odds ratios and Shapley value decomposition, GSS 2022–2024"
  ) %>%
  fmt_number(columns = OR, decimals = 2) %>%
  fmt_number(columns = Percent_of_Fit, decimals = 1) %>%
  fmt_integer(columns = Rank) %>%
  cols_label(
    variable       = "Predictor",
    OR             = "OR (geom. mean)",
    `95% CI`       = "95% CI",
    Percent_of_Fit = "% of Fit (Shapley)",
    Rank           = "Rank"
  ) %>%
  tab_source_note(
    "OR = geometric mean of exponentiated coefficients within factor variable. CI = [min, max] across contrasts. Shapley contribution = average marginal R\u00b2 across all 2^k subset models (McFadden)."
  ) %>%
  tab_options(table.font.size = 11)

gt_table1

gtsave(
  gt_table1,
  file.path(out_tables, "logistic_shapley_predictors.png"),
  expand = 10
)

# Export table1_data as CSV for paper
write.csv(
  table1_data %>% select(variable, OR, CI_low, CI_high, Percent_of_Fit, Rank),
  file.path(out_data, "logistic_shapley_table1.csv"),
  row.names = FALSE
)
```

------------------------------------------------------------------------

## Section 2: Mode Gaps in Substantive Responses (§2.2)

This section estimates the association between survey mode and each binary outcome variable, comparing bivariate mode gaps to gaps after demographic adjustment. Both models are weighted by `wtssps`.

### Outcome Variable Definitions

```{r define-formulas}
# Analytic core variable set — defined in 00_setup.R, sourced from
# data/analytic_core.csv (analytic_core == 1). Filtered to variables
# actually present in the analysis dataset.
c_vars <- intersect(ANALYTIC_CORE_VARS, names(analysis_data))

# Model formula strings
bivariate_formula_part <- "mode_fctr"
controls_formula_part  <- "mode_fctr + year + agegrp + degree_cat + marital_cat + w_hisp_race + f_born + f_sex + f_region"
```

### Bivariate Model Loop

```{r run-bivar, warning=FALSE}
all_model_results_bivariate <- list()

for (var in c_vars) {

  if (!var %in% names(analysis_data)) next
  analysis_data[[var]] <- as.numeric(as.character(analysis_data[[var]]))
  unique_vals <- unique(analysis_data[[var]][!is.na(analysis_data[[var]])])
  if (length(unique_vals) < 2) next
  if (!all(unique_vals %in% c(0, 1))) next

  model_fit <- tryCatch(
    glm(
      as.formula(paste(var, "~", bivariate_formula_part)),
      family  = binomial(link = "logit"),
      data    = analysis_data,
      weights = wtssps
    ),
    error = function(e) NULL
  )

  if (!is.null(model_fit)) {
    tryCatch({
      coef_df <- as.data.frame(summary(model_fit)$coefficients) %>%
        rownames_to_column("term")
      ci_df   <- as.data.frame(confint.default(model_fit)) %>%
        rownames_to_column("term")
      names(coef_df) <- c("term", "estimate", "std.error", "statistic", "p.value")
      names(ci_df)   <- c("term", "conf.low", "conf.high")

      all_model_results_bivariate[[var]] <- left_join(coef_df, ci_df, by = "term") %>%
        mutate(variable = var, model_type = "Bivariate") %>%
        select(variable, model_type, term, estimate, std.error, statistic, p.value, conf.low, conf.high)
    }, error = function(e) NULL)
  }
}

final_model_results_bivariate <- bind_rows(all_model_results_bivariate) %>%
  group_by(term) %>%
  mutate(
    p.value.adj = if_else(
      term == "mode_fctrweb",
      p.adjust(p.value, method = "holm"),
      NA_real_
    )
  ) %>%
  ungroup()
```

### Full Model Loop

```{r run-full, warning=FALSE}
all_model_results_full <- list()

for (var in c_vars) {

  if (!var %in% names(analysis_data)) next
  analysis_data[[var]] <- as.numeric(as.character(analysis_data[[var]]))
  unique_vals <- unique(analysis_data[[var]][!is.na(analysis_data[[var]])])
  if (length(unique_vals) < 2) next
  if (!all(unique_vals %in% c(0, 1))) next

  model_fit <- tryCatch(
    glm(
      as.formula(paste(var, "~", controls_formula_part)),
      family  = binomial(link = "logit"),
      data    = analysis_data,
      weights = wtssps
    ),
    error = function(e) NULL
  )

  if (!is.null(model_fit)) {
    tryCatch({
      coef_df <- as.data.frame(summary(model_fit)$coefficients) %>%
        rownames_to_column("term")
      ci_df   <- as.data.frame(confint.default(model_fit)) %>%
        rownames_to_column("term")
      names(coef_df) <- c("term", "estimate", "std.error", "statistic", "p.value")
      names(ci_df)   <- c("term", "conf.low", "conf.high")

      all_model_results_full[[var]] <- left_join(coef_df, ci_df, by = "term") %>%
        mutate(variable = var, model_type = "Full") %>%
        select(variable, model_type, term, estimate, std.error, statistic, p.value, conf.low, conf.high)
    }, error = function(e) NULL)
  }
}

final_model_results_full <- bind_rows(all_model_results_full) %>%
  group_by(term) %>%
  mutate(
    p.value.adj = if_else(
      term == "mode_fctrweb",
      p.adjust(p.value, method = "holm"),
      NA_real_
    )
  ) %>%
  ungroup()
```

### Significance Summaries

```{r summary-bivar}
p_vals_bivariate <- final_model_results_bivariate %>%
  mutate(sig.check = !is.na(p.value) & p.value < 0.05) %>%
  select(variable, term, sig.check)

sig.chart_bivariate <- p_vals_bivariate %>%
  group_by(term) %>%
  summarise(
    significant_count = sum(sig.check, na.rm = TRUE),
    total_models      = n(),
    sig.prop          = significant_count / total_models,
    .groups = "drop"
  ) %>%
  arrange(desc(sig.prop))

sig.chart_bivariate %>%
  gt() %>%
  tab_header(title = "Term significance (p < 0.05) — Bivariate model (wtssps)") %>%
  fmt_number(columns = sig.prop, decimals = 3) %>%
  cols_label(
    term              = "Term",
    significant_count = "Sig Count",
    total_models      = "Total Models",
    sig.prop          = "Sig Proportion"
  )
```

```{r summary-full}
p_vals_full <- final_model_results_full %>%
  mutate(
    sig.check     = !is.na(p.value)     & p.value     < 0.05,
    sig.check.adj = !is.na(p.value.adj) & p.value.adj < 0.05
  ) %>%
  select(variable, term, sig.check, sig.check.adj, p.value, p.value.adj)

sig.chart_full <- p_vals_full %>%
  group_by(term) %>%
  summarise(
    significant_count      = sum(sig.check, na.rm = TRUE),
    significant_count_holm = sum(sig.check.adj, na.rm = TRUE),
    total_models           = n(),
    sig.prop               = significant_count / total_models,
    sig.prop.holm          = significant_count_holm / sum(!is.na(sig.check.adj)),
    .groups = "drop"
  ) %>%
  arrange(desc(sig.prop))

sig.chart_full %>%
  gt() %>%
  tab_header(title = "Term significance (p < 0.05) — Full model (wtssps), with Holm correction for mode_fctrweb") %>%
  fmt_number(columns = c(sig.prop, sig.prop.holm), decimals = 3) %>%
  cols_label(
    term                   = "Term",
    significant_count      = "Sig Count",
    significant_count_holm = "Sig Count (Holm)",
    total_models           = "Total Models",
    sig.prop               = "Sig Proportion",
    sig.prop.holm          = "Sig Proportion (Holm)"
  )
```

```{r mode-effect-detail}
final_model_results_full %>%
  filter(term == "mode_fctrweb") %>%
  arrange(p.value) %>%
  mutate(
    sig_uncorrected = if_else(p.value < 0.05, "*", ""),
    sig_holm        = if_else(!is.na(p.value.adj) & p.value.adj < 0.05, "*", "")
  ) %>%
  select(variable, estimate, std.error, statistic, p.value, p.value.adj,
         sig_uncorrected, sig_holm) %>%
  head(20) %>%
  gt() %>%
  tab_header(title = "Top 20 mode effects — Full model (wtssps)") %>%
  fmt_number(columns = c(estimate, std.error, statistic, p.value, p.value.adj), decimals = 4) %>%
  cols_label(
    variable        = "Variable",
    estimate        = "Estimate",
    std.error       = "SE",
    statistic       = "Statistic",
    p.value         = "p-value",
    p.value.adj     = "p-value (Holm)",
    sig_uncorrected = "Sig (α=0.05)",
    sig_holm        = "Sig (Holm)"
  )
```

```{r compare-sig}
comparison <- p_vals_bivariate %>%
  rename(bivar.sig = sig.check) %>%
  left_join(
    p_vals_full %>% select(variable, term, sig.check) %>% rename(full.sig = sig.check),
    by = join_by(variable, term)
  ) %>%
  filter(term == "mode_fctrweb") %>%
  mutate(
    pattern = case_when(
      bivar.sig  &  full.sig ~ "Significant in both (measurement effect)",
      bivar.sig  & !full.sig ~ "Bivariate only (compositional)",
      !bivar.sig &  full.sig ~ "Full only (suppressed bivariate)",
      TRUE                   ~ "Non-significant in both"
    )
  ) %>%
  select(variable, bivar.sig, full.sig, pattern) %>%
  arrange(pattern, variable)

# Pattern summary
comparison %>%
  count(pattern) %>%
  mutate(prop = round(n / sum(n), 3)) %>%
  gt() %>%
  tab_header(title = "Mode effect pattern summary (mode_fctrweb)")

# Variable-level table (excluding non-significant in both)
comparison %>%
  filter(pattern != "Non-significant in both") %>%
  gt() %>%
  tab_header(title = "Variables with mode effect (mode_fctrweb) — excluding non-significant in both") %>%
  cols_label(
    variable  = "Variable",
    bivar.sig = "Sig Bivariate",
    full.sig  = "Sig Full",
    pattern   = "Pattern"
  )
```

### Diagnostic Summaries

```{r figure6-diagnostics}
# Construct mode_gaps_comparison for diagnostics and Figure 3
bivariate_mode_effects <- final_model_results_bivariate %>%
  filter(term == "mode_fctrweb") %>%
  select(variable,
         estimate_bivariate = estimate,
         se_bivariate       = std.error,
         p_bivariate        = p.value,
         p_bivariate_adj    = p.value.adj)

full_mode_effects <- final_model_results_full %>%
  filter(term == "mode_fctrweb") %>%
  select(variable,
         estimate_full = estimate,
         se_full       = std.error,
         p_full        = p.value,
         p_full_adj    = p.value.adj)

mode_gaps_comparison <- bivariate_mode_effects %>%
  inner_join(full_mode_effects, by = "variable") %>%
  mutate(
    sig_bivariate = !is.na(p_bivariate_adj) & p_bivariate_adj < 0.05,
    sig_full      = !is.na(p_full_adj)      & p_full_adj      < 0.05,
    significance_pattern = case_when(
      sig_bivariate &  sig_full ~ "Significant (Both)",
      sig_bivariate & !sig_full ~ "Reduced by Controls",
     !sig_bivariate &  sig_full ~ "Emerged with Controls",
      TRUE                      ~ "Not Significant"
    ),
    gap_change    = estimate_full - estimate_bivariate,
    gap_reduction = abs(estimate_bivariate) - abs(estimate_full),
    var_display   = gsub("^c_", "", variable) %>% toupper()
  )

# Top 10 persistent gaps (cited §2.2)
persistent_gaps <- mode_gaps_comparison %>%
  filter(sig_full) %>%
  arrange(desc(abs(estimate_full))) %>%
  head(10) %>%
  select(var_display, estimate_bivariate, estimate_full, gap_reduction, significance_pattern)

persistent_gaps %>%
  gt() %>%
  tab_header(title = "Top 10 variables with largest persistent mode gaps after demographic adjustment") %>%
  fmt_number(columns = where(is.numeric), decimals = 4)

# Variables where adjustment eliminated gaps (cited §2.2)
eliminated_gaps <- mode_gaps_comparison %>%
  filter(significance_pattern == "Reduced by Controls") %>%
  arrange(desc(gap_reduction)) %>%
  head(10) %>%
  select(var_display, estimate_bivariate, estimate_full, gap_reduction)

eliminated_gaps %>%
  gt() %>%
  tab_header(title = "Top 10 variables where demographic adjustment eliminated mode gaps") %>%
  fmt_number(columns = where(is.numeric), decimals = 4)
```

### Mode Gaps Before and After Demographic Adjustment

```{r mode-gaps-scatter, fig.width=10, fig.height=10}
# Significance pattern summary
mode_gaps_comparison %>%
  count(significance_pattern) %>%
  mutate(pct = round(n / sum(n) * 100, 1)) %>%
  gt() %>%
  tab_header(title = paste0("Mode gap significance patterns — ", nrow(mode_gaps_comparison), " variables")) %>%
  cols_label(significance_pattern = "Significance Pattern", n = "N", pct = "%")

mode_gaps_plot <- mode_gaps_comparison %>%
  filter(variable != "c_discaffw") %>% 
  mutate(
    plot_x = if_else(estimate_bivariate < 0, -estimate_bivariate, estimate_bivariate),
    plot_y = if_else(estimate_full < 0, -estimate_full,      estimate_full)
  )


fig3 <- ggplot(
  mode_gaps_plot,
  aes(x = plot_x, y = plot_y,
      color = significance_pattern, shape = significance_pattern)
) +
  geom_abline(intercept = 0, slope = 1,
              linetype = "dashed", color = "gray50", linewidth = 0.8) +
  geom_hline(yintercept = 0,
             linetype = "solid", color = "gray30", linewidth = 0.5) +
  geom_vline(xintercept = 0,
             linetype = "solid", color = "gray30", linewidth = 0.5) +
  geom_point(size = 3.5, alpha = 0.7) +
  scale_color_manual(
    values = c(
      "Significant (Both)"    = "#D95F02",
      "Reduced by Controls"   = "#1B9E77",
      "Emerged with Controls" = "#7570B3",
      "Not Significant"       = "#999999"
    ),
    name = "Significance Pattern"
  ) +
  scale_shape_manual(
    values = c(
      "Significant (Both)"    = 16,
      "Reduced by Controls"   = 17,
      "Emerged with Controls" = 15,
      "Not Significant"       = 1
    ),
    name = "Significance Pattern"
  ) +
  labs(
    x        = "Raw Mode Gap\n(Bivariate Model Coefficient)",
    y        = "Adjusted Mode Gap\n(Full Model Coefficient with Demographic Controls)",
    caption  = paste0(
      "Points on diagonal = adjustment has no effect. Below diagonal = gap reduced by controls. ",
      "Above diagonal = gap increased by controls.\nN = ", nrow(mode_gaps_comparison), " variables."
    )
  ) +
  coord_equal() +
  annotate(
    "text",
    x     = min(mode_gaps_plot$plot_x) * 0.8,
    y     = max(mode_gaps_plot$plot_y) * 0.9,
    label = "Diagonal = No effect of adjustment\nAbove diagonal = Gap increases with controls\nBelow diagonal = Gap decreases with controls",
    hjust = 0, size = 3, color = "gray30", fontface = "italic"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position  = "bottom",
    legend.title     = element_text(face = "bold", size = 13),
    legend.text      = element_text(size = 12),
    panel.grid.minor = element_blank(),
    axis.title       = element_text(size = 13, face = "bold"),
    axis.text        = element_text(size = 11),
    plot.caption     = element_text(size = 10)
  )

fig3

ggsave(
  filename = file.path(out_fig, "mode_gaps_scatter.png"),
  plot     = fig3,
  width    = 10, height = 10, dpi = 300
)
```

------------------------------------------------------------------------

## Save Results

```{r save-results}
write_csv(mode_gaps_comparison,
          file.path(out_data, "mode_gaps_comparison.csv"))
write_csv(final_model_results_bivariate,
          file.path(out_data, "gss_glm_mode_bivariate_22_24.csv"))
write_csv(final_model_results_full,
          file.path(out_data, "gss_glm_mode_with_controls_22_24.csv"))
```
