---
title: "GSS Mode Effects: Temporal Persistence Analysis (2016–2024)"
format:
  html:
    toc-depth: 3
editor: visual
---

## Overview

This file produces all §4 outputs examining which survey mode better preserves temporal continuity with the pre-multimode GSS baseline (2016–2018):

-   **Forest plots** comparing 2016–2018 baseline estimates to 2022–2024 mode-specific estimates with 84% confidence intervals.
-   **Violin plot** showing distribution of absolute changes across all 137 variables for each mode.
-   **Per-domain forest plots** facilitating visual inspection of temporal persistence by thematic area.

Four comparison groups are used: 2016–2018 pooled baseline (gray), 2022–2024 full combined sample (black), 2022–2024 FTF/Phone subsample (orange), and 2022–2024 Web subsample (teal).

------------------------------------------------------------------------

## Setup

```{r setup, message=FALSE, warning=FALSE}
source("00_setup.R")

# File-specific libraries not covered by 00_setup.R
library(gt)
library(forcats)

# Output directories
out_fig    <- "outputs/figures"
out_tables <- "outputs/tables"
out_data   <- "outputs/data"
for (d in c(out_fig, out_tables, out_data)) {
  dir.create(d, recursive = TRUE, showWarnings = FALSE)
}

# Define all possible groups (will be filtered below based on available designs)
all_group_levels <- c(
  "2016-2018 (all)",
  "2022-2024 (all)",
  "2022-2024 (f2fp)",
  "2022-2024 (web)"
)

all_color_mapping <- c(
  "2016-2018 (all)"  = unname(MODE_COLORS["2016-2018"]),
  "2022-2024 (all)"  = unname(MODE_COLORS["Population"]),
  "2022-2024 (f2fp)" = unname(MODE_COLORS["f2fp"]),
  "2022-2024 (web)"  = unname(MODE_COLORS["web"])
)

all_legend_order <- c(
  "2016-2018 (all)", "2022-2024 (all)",
  "2022-2024 (f2fp)", "2022-2024 (web)"
)

domain_order <- c(
  "Religion",
  "Race & Inequality",
  "Civil Liberties & Government",
  "Crime, Punishment & Firearms",
  "Political & Economic Orientations",
  "Social Attitudes & Morality",
  "Institutional Confidence",
  "Well-being & Life Orientations"
)
```

------------------------------------------------------------------------

## Data Loading

```{r load-data, message=FALSE, warning=FALSE}
if (!file.exists(GSS_RDS_PATH))  stop("RDS not found: ", GSS_RDS_PATH)
if (!file.exists(WEIGHTS_PATH))  stop("Weights not found: ", WEIGHTS_PATH)

gss_data     <- readRDS(GSS_RDS_PATH)
mode_weights <- read_dta(WEIGHTS_PATH)

# 2022-2024 analysis dataset: join mode-specific weights, drop mode==3
gss_analysis_data <- gss_data %>%
  filter(year %in% c(2022, 2024), mode != 3) %>%
  left_join(mode_weights, by = join_by(id, year)) %>%
  mutate(
    wtssps_f2fp = if_else(is.na(wtssps_f2fp), wtssps, wtssps_f2fp),
    wtssps_web  = if_else(is.na(wtssps_web),  wtssps, wtssps_web)
  ) %>%
  filter(!is.na(vpsu), !is.na(vstrat), !is.na(wtssps), !is.na(mode)) %>%
  mutate(year = as.numeric(as.character(year)))

```

------------------------------------------------------------------------

## Analytic Core Variables

```{r define-core-vars}
# Analytic core variable set — defined in 00_setup.R, sourced from
# data/analytic_core.csv (analytic_core == 1). All 137 items are fielded in
# 2016, 2018, 2022, and 2024; the CSV is the authoritative record of inclusion.
c_vars <- intersect(ANALYTIC_CORE_VARS, names(gss_data))

```

------------------------------------------------------------------------

## Survey Design Objects — 4 Groups

```{r define-designs}
options(survey.lonely.psu = "adjust")

safe_svydesign <- function(ids, strata, weights, data) {
  tryCatch(
    svydesign(ids = ids, strata = strata, weights = weights, data = data, nest = TRUE),
    error = function(e) { message("Error creating design: ", e$message); NULL }
  )
}

# 2016-2018 pooled baseline
data_1618_pooled   <- gss_data %>%
  filter(year %in% c(2016, 2018), !is.na(vpsu), !is.na(vstrat), !is.na(wtssps))
design_1618_pooled <- safe_svydesign(~vpsu, ~vstrat, ~wtssps, data_1618_pooled)

# 2022-2024: all modes combined
design_all_pooled  <- safe_svydesign(~vpsu, ~vstrat, ~wtssps, gss_analysis_data)

# 2022-2024: FTF/Phone only (mode-specific weight)
data_f2fp_pooled   <- subset(gss_analysis_data, mode %in% c(1, 2) & !is.na(wtssps_f2fp))
design_f2fp_pooled <- if (nrow(data_f2fp_pooled) > 0)
  safe_svydesign(~vpsu, ~vstrat, ~wtssps_f2fp, data_f2fp_pooled) else NULL

# 2022-2024: Web only (mode-specific weight)
data_web_pooled    <- subset(gss_analysis_data, mode == 4 & !is.na(wtssps_web))
design_web_pooled  <- if (nrow(data_web_pooled) > 0)
  safe_svydesign(~vpsu, ~vstrat, ~wtssps_web, data_web_pooled) else NULL

if (is.null(design_all_pooled) || is.null(design_1618_pooled)) {
  stop("Failed to create main survey design objects. Check data.")
}
```

------------------------------------------------------------------------

## Calculate 84% Confidence Intervals

84% CIs are used because non-overlapping 84% intervals correspond approximately to a significant pairwise difference at *p* \< .05 (Payton, Greenstone, & Schenker, 2003).

```{r calc-cis}
calculate_ci <- function(design_obj, var_formula) {
  if (is.null(design_obj)) return(list(est = NA, se = NA, lwr = NA, upr = NA))
  mean_obj <- tryCatch(svymean(var_formula, design_obj, na.rm = TRUE),
                       error = function(e) NA)
  if (any(is.na(mean_obj)) || !inherits(mean_obj, "svystat"))
    return(list(est = NA, se = NA, lwr = NA, upr = NA))
  est <- tryCatch(as.numeric(coef(mean_obj)[1]), error = function(e) NA)
  se  <- tryCatch(as.numeric(SE(mean_obj)[1]),   error = function(e) NA)
  if (is.na(est) || is.na(se) || se == 0)
    return(list(est = NA, se = NA, lwr = NA, upr = NA))
  ci_obj <- tryCatch(confint(mean_obj, level = 0.84), error = function(e) NA)
  if (any(is.na(ci_obj)) || !is.matrix(ci_obj) || nrow(ci_obj) != 1 || ncol(ci_obj) != 2) {
    ci_lwr <- NA; ci_upr <- NA
  } else {
    ci_lwr <- ci_obj[1, 1]; ci_upr <- ci_obj[1, 2]
  }
  list(
    est = est, se = se,
    lwr = ifelse(is.na(ci_lwr), NA, max(0, ci_lwr)),
    upr = ifelse(is.na(ci_upr), NA, min(1, ci_upr))
  )
}

# Build design list conditionally — only include groups with valid designs
design_list <- list(
  "2016-2018 (all)"  = list(d = design_1618_pooled, data = data_1618_pooled),
  "2022-2024 (all)"  = list(d = design_all_pooled,  data = gss_analysis_data)
)

# Add mode-specific designs only if they exist
if (!is.null(design_f2fp_pooled)) {
  design_list[["2022-2024 (f2fp)"]] <- list(d = design_f2fp_pooled, data = data_f2fp_pooled)
}
if (!is.null(design_web_pooled)) {
  design_list[["2022-2024 (web)"]] <- list(d = design_web_pooled, data = data_web_pooled)
}

# Filter aesthetics to only include groups with valid designs
actual_groups <- names(design_list)
group_levels  <- all_group_levels[all_group_levels %in% actual_groups]
color_mapping <- all_color_mapping[names(all_color_mapping) %in% actual_groups]
legend_order  <- all_legend_order[all_legend_order %in% actual_groups]

ci_results_pooled <- data.frame(
  variable   = character(),
  mode_group = character(),
  estimate   = numeric(),
  se         = numeric(),
  ci_lower   = numeric(),
  ci_upper   = numeric(),
  stringsAsFactors = FALSE
)


for (var in c_vars) {
  str_formula <- as.formula(paste0("~", var))
  row_offset  <- nrow(ci_results_pooled)
  for (group_name in names(design_list)) {
    d_info <- design_list[[group_name]]
    if (!is.null(d_info$data) && var %in% names(d_info$data)) {
      res <- calculate_ci(d_info$d, str_formula)
      ci_results_pooled[row_offset + 1, ] <-
        list(var, group_name, res$est, res$se, res$lwr, res$upr)
    } else {
      ci_results_pooled[row_offset + 1, ] <- list(var, group_name, NA, NA, NA, NA)
    }
    row_offset <- row_offset + 1
  }
}

```

------------------------------------------------------------------------

## Topic Mapping and Domain Consolidation

```{r topic-mapping}
# Read domain and topic assignments directly from the analytic core registry.
# The eight final domains and variable-level overrides (e.g., CAPPUN, DISCAFF,
# LETIN1A) are already incorporated in data/gss24_attitudes_opinions.csv.
topic_mapping <- readr::read_csv("data/gss24_attitudes_opinions.csv", show_col_types = FALSE) |>
  dplyr::filter(analytic_core == 1) |>
  dplyr::transmute(
    varname  = var,
    variable = paste0("c_", tolower(var)),
    topic    = category,
    domain   = factor(domain, levels = domain_order)
  )

# Variable-domain lookup saved for reference
domain_var_table <- topic_mapping %>%
  mutate(var_display = toupper(gsub("^c_", "", variable))) %>%
  arrange(domain, var_display) %>%
  select(Domain = domain, Variable = var_display, `Source Topic` = topic)

write_csv(domain_var_table, file.path(out_data, "domain_variable_lookup.csv"))

domain_var_table %>%
  count(Domain, name = "N") %>%
  gt() %>%
  tab_header(title = "Variables per domain")
```

------------------------------------------------------------------------

## Helper: Plot Theme

```{r theme-helper}
# Centralised theme so tweaks propagate to all plots
forest_theme <- function(base_size = 11) {
  theme_bw(base_size = base_size) +
    theme(
      plot.title        = element_text(hjust = 0.5, face = "bold", size = base_size + 3),
      plot.subtitle     = element_text(hjust = 0.5, size = base_size),
      legend.position   = "bottom",
      legend.text       = element_text(size = base_size - 2),
      legend.title      = element_text(face = "bold"),
      strip.placement   = "outside",
      strip.background  = element_blank(),
      strip.text.y.left = element_text(angle = 0, face = "bold", hjust = 1),
      panel.grid.minor  = element_blank()
    )
}
```

------------------------------------------------------------------------

## Forest Plot — All Variables (Appendix)

### Full Forest Plot Across All Variables

```{r forest-plot-all, fig.width=12, fig.height=20}
plot_data_all <- ci_results_pooled %>%
  filter(!is.na(estimate), !is.na(ci_lower), !is.na(ci_upper)) %>%
  left_join(topic_mapping, by = "variable") %>%
  mutate(
    domain     = replace_na(as.character(domain), "Other / Uncategorized"),
    domain     = factor(domain, levels = domain_order),
    mode_group = factor(mode_group, levels = group_levels),
    variable   = fct_rev(factor(toupper(gsub("^c_", "", variable))))
  )

if (nrow(plot_data_all) > 0) {
  forest_all <- ggplot(plot_data_all,
                       aes(x = estimate, y = variable, color = mode_group)) +
    geom_point(position = position_dodge(width = 0.7), size = 2) +
    geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper),
                   position = position_dodge(width = 0.7),
                   height = 0, linewidth = 0.7) +
    facet_grid(domain ~ ., scales = "free_y", space = "free_y", switch = "y") +
    scale_color_manual(values = color_mapping, breaks = legend_order, name = "Group") +
    guides(color = guide_legend(nrow = 2, byrow = TRUE)) +
    labs(
      title    = "84% Confidence Intervals by Domain and Group",
      subtitle = "GSS 2016–18 baseline vs. 2022–24, by mode",
      x        = "Estimated Proportion",
      y        = NULL
    ) +
    forest_theme() +
    coord_cartesian(xlim = c(0, 1))

  forest_all

  plot_height <- max(8, n_distinct(plot_data_all$variable) * 0.5)
  ggsave(
    file.path(out_fig, "forest_all_variables.png"),
    plot = forest_all, width = 12, height = plot_height,
    units = "in", dpi = 300, limitsize = FALSE
  )

  # --- Per-domain forest plots ---
  plot_domain <- function(domain_name, data) {
    domain_data <- data %>% filter(as.character(domain) == domain_name)
    n_vars      <- n_distinct(domain_data$variable)
    p <- ggplot(domain_data,
                aes(x = estimate, y = variable, color = mode_group)) +
      geom_point(position = position_dodge(width = 0.7), size = 2.5) +
      geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper),
                     position = position_dodge(width = 0.7),
                     height = 0, linewidth = 0.8) +
      scale_color_manual(values = color_mapping, breaks = legend_order, name = "Group") +
      guides(color = guide_legend(nrow = 2, byrow = TRUE)) +
      labs(
        title    = domain_name,
        subtitle = "84% CIs — GSS 2016–18 baseline vs. 2022–24, by mode",
        x        = "Estimated Proportion",
        y        = NULL
      ) +
      forest_theme() +
      coord_cartesian(xlim = c(0, 1))
    list(plot = p, height = max(4, n_vars * 0.45))
  }

  safe_name <- function(x) {
    tolower(x) %>%
      str_replace_all("[^a-z0-9]+", "_") %>%
      str_remove("_$")
  }

  out_fig_domains <- file.path(out_fig, "by_domain")
  dir.create(out_fig_domains, recursive = TRUE, showWarnings = FALSE)

  domains_present <- levels(droplevels(plot_data_all$domain))
  for (dom in domains_present) {
    result <- plot_domain(dom, plot_data_all)
    fname  <- paste0("forest_", safe_name(dom), ".png")
    ggsave(
      file.path(out_fig_domains, fname),
      plot = result$plot, width = 10, height = result$height,
      units = "in", dpi = 300, limitsize = FALSE
    )
  }
}
```

```{r save-plot-data}
# Save plot_data_all to CSV so appendix_forest_plots.qmd can reproduce all
# forest plots without re-running the survey design and CI calculation pipeline.
write_csv(
  plot_data_all %>%
    mutate(
      variable   = as.character(variable),
      domain     = as.character(domain),
      mode_group = as.character(mode_group)
    ),
  file.path(out_data, "forest_plot_data.csv")
)
```

------------------------------------------------------------------------

## Forest Plot — Racial Attitudes Domain

### Racial Attitudes Domain Forest Plot

```{r forest-plot-racial, fig.width=12, fig.height=10}
racial_topics <- unique(plot_data_all$domain)[
  str_detect(tolower(as.character(unique(plot_data_all$domain))), "racial|race")
]

plot_data_racial <- plot_data_all %>%
  filter(domain %in% racial_topics)

if (nrow(plot_data_racial) > 0) {
  forest_racial <- ggplot(plot_data_racial,
                          aes(x = estimate, y = variable, color = mode_group)) +
    geom_point(position = position_dodge(width = 0.7), size = 2.5) +
    geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper),
                   position = position_dodge(width = 0.7),
                   height = 0, linewidth = 0.8) +
    facet_grid(domain ~ ., scales = "free_y", space = "free_y", switch = "y") +
    scale_color_manual(values = color_mapping, breaks = legend_order, name = "Group") +
    guides(color = guide_legend(nrow = 2, byrow = TRUE)) +
    labs(
      title    = "84% Confidence Intervals — Racial Attitudes Domain",
      subtitle = "GSS 2016–18 baseline vs. 2022–24, by mode",
      x        = "Estimated Proportion",
      y        = NULL,
      caption  = paste0(
        "84% non-overlapping CIs correspond approximately to p < .05 pairwise difference.\n",
        "FTF/Phone (orange); Web (teal); Full sample (black); Baseline 2016–18 (gray)."
      )
    ) +
    forest_theme() +
    coord_cartesian(xlim = c(0, 1))

  forest_racial

  plot_height_racial <- max(6, n_distinct(plot_data_racial$variable) * 0.5)
  ggsave(
    file.path(out_fig, "forest_racial_attitudes.png"),
    plot = forest_racial, width = 12, height = plot_height_racial,
    units = "in", dpi = 300, limitsize = FALSE
  )
} else {
  warning("No variables matched racial attitudes filter. Check domain labels above.")
}
```

------------------------------------------------------------------------

## Closeness Analysis

```{r closeness-analysis}
closeness_results <- ci_results_pooled %>%
  select(variable, mode_group, estimate) %>%
  filter(!is.na(estimate)) %>%
  pivot_wider(names_from = mode_group, values_from = estimate) %>%
  select(
    variable,
    est_1618 = `2016-2018 (all)`,
    est_all  = `2022-2024 (all)`,
    est_f2fp = `2022-2024 (f2fp)`,
    est_web  = `2022-2024 (web)`
  ) %>%
  mutate(
    dist_to_f2fp = abs(est_1618 - est_f2fp),
    dist_to_web  = abs(est_1618 - est_web),
    dist_to_all  = abs(est_1618 - est_all),
    closer_to    = case_when(
      dist_to_f2fp < dist_to_web  ~ "f2fp",
      dist_to_web  < dist_to_f2fp ~ "web",
      dist_to_web == dist_to_f2fp ~ "tie",
      TRUE                        ~ NA_character_
    ),
    # Is FTF closer to baseline than the full combined sample?
    ftf_beats_full = dist_to_f2fp < dist_to_all
  )

n_total          <- nrow(closeness_results)
n_ftf_beats_web  <- sum(closeness_results$closer_to == "f2fp", na.rm = TRUE)
n_web_beats_ftf  <- sum(closeness_results$closer_to == "web",  na.rm = TRUE)
n_ftf_beats_full <- sum(closeness_results$ftf_beats_full,      na.rm = TRUE)
n_web_beats_full <- sum(closeness_results$dist_to_web < closeness_results$dist_to_all, na.rm = TRUE)

summary_table <- tibble(
  Mode            = c("FTF/Phone", "Web", "Full Sample"),
  `N closer to baseline (vs other mode)` = c(
    paste0(n_ftf_beats_web,  " (", round(n_ftf_beats_web  / n_total * 100, 1), "%)"),
    paste0(n_web_beats_ftf,  " (", round(n_web_beats_ftf  / n_total * 100, 1), "%)"),
    "—"
  ),
  `N closer to baseline (vs full sample)` = c(
    paste0(n_ftf_beats_full, " (", round(n_ftf_beats_full / n_total * 100, 1), "%)"),
    paste0(n_web_beats_full, " (", round(n_web_beats_full / n_total * 100, 1), "%)"),
    "reference"
  ),
  `Mean |distance|` = c(
    round(mean(closeness_results$dist_to_f2fp, na.rm = TRUE), 3),
    round(mean(closeness_results$dist_to_web,  na.rm = TRUE), 3),
    round(mean(closeness_results$dist_to_all,  na.rm = TRUE), 3)
  ),
  `Median |distance|` = c(
    round(median(closeness_results$dist_to_f2fp, na.rm = TRUE), 3),
    round(median(closeness_results$dist_to_web,  na.rm = TRUE), 3),
    round(median(closeness_results$dist_to_all,  na.rm = TRUE), 3)
  ),
  `Max |distance|` = c(
    round(max(closeness_results$dist_to_f2fp, na.rm = TRUE), 3),
    round(max(closeness_results$dist_to_web,  na.rm = TRUE), 3),
    round(max(closeness_results$dist_to_all,  na.rm = TRUE), 3)
  )
)

summary_table %>%
  gt() %>%
  tab_header(
    title    = "Baseline proximity summary",
    subtitle = paste0(
      "Distance from 2016–2018 estimates across ", n_total, " variables. ",
      "'Closer to baseline (vs other mode)' = FTF vs Web head-to-head. ",
      "'Closer to baseline (vs full sample)' = mode-specific vs pooled 2022–2024 estimate."
    )
  )

write_csv(closeness_results, file.path(out_data, "closeness_by_variable.csv"))
```

------------------------------------------------------------------------

## Distance Distribution Plots

### Violin Plot of Baseline Distance Distributions

```{r distance-plots, fig.width=12, fig.height=7}
# Pivot to long including full sample
closeness_long <- closeness_results %>%
  filter(!is.na(dist_to_f2fp), !is.na(dist_to_web), !is.na(dist_to_all)) %>%
  pivot_longer(
    cols      = c(dist_to_f2fp, dist_to_web, dist_to_all),
    names_to  = "mode",
    values_to = "distance"
  ) %>%
  mutate(
    mode_label = case_when(
      mode == "dist_to_f2fp" ~ "FTF/Phone",
      mode == "dist_to_web"  ~ "Web",
      mode == "dist_to_all"  ~ "Full Sample"
    ),
    mode_label = factor(mode_label, levels = c("FTF/Phone", "Full Sample", "Web"))
  )

distance_stats <- closeness_long %>%
  group_by(mode_label) %>%
  summarise(
    median = median(distance, na.rm = TRUE),
    mean   = mean(distance, na.rm = TRUE),
    n      = n(),
    .groups = "drop"
  ) %>%
  arrange(mode_label)

# Identify outliers (beyond 1.5 * IQR) per mode group
outlier_data <- closeness_long %>%
  group_by(mode_label) %>%
  mutate(
    q1         = quantile(distance, 0.25, na.rm = TRUE),
    q3         = quantile(distance, 0.75, na.rm = TRUE),
    iqr        = q3 - q1,
    is_outlier = distance > (q3 + 1.5 * iqr)   # only upper outliers (distance >= 0)
  ) %>%
  filter(is_outlier) %>%
  mutate(var_label = toupper(gsub("^c_", "", variable))) %>%
  ungroup()

# x-axis nudge direction: FTF labels left, Full Sample right, Web right
outlier_data <- outlier_data %>%
  mutate(
    hjust_val = if_else(mode_label == "FTF/Phone", 1.15, -0.15),
    nudge_x   = if_else(mode_label == "FTF/Phone", -0.05, 0.05)
  )

violin_plot <- ggplot(closeness_long,
                      aes(x = mode_label, y = distance)) +
  geom_violin(alpha = 0.55, trim = TRUE, fill = "gray90") +
  geom_boxplot(width = 0.15, alpha = 0.85, outlier.shape = NA,
               color = "gray20") +
  # Outlier dots
  geom_point(data = outlier_data,
             aes(x = mode_label, y = distance),
             size = 2, inherit.aes = FALSE) +
  # Outlier labels — offset horizontally away from violin
  geom_text(data = outlier_data,
            aes(x    = as.numeric(mode_label) + nudge_x,
                y    = distance,
                label = var_label,
                hjust = hjust_val),
            size = 2.5, color = "gray25", inherit.aes = FALSE) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3,
               fill = "white", color = "black") +
  scale_y_continuous(
    labels = scales::number_format(accuracy = 0.01),
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.08))
  ) +
  # Stats annotation below each violin
  geom_text(
    data = distance_stats,
    aes(x = mode_label,
        y = -Inf,
        label = sprintf("Median: %.3f\nMean: %.3f", median, mean)),
    vjust = -0.3, size = 3, color = "gray30", inherit.aes = FALSE
  ) +
  labs(
    title   = "Distribution of Distance from 2016\u20132018 Baseline",
    subtitle = "Absolute difference |p_baseline \u2212 p_mode| across all variables",
    x       = NULL,
    y       = "Distance from Baseline (Absolute Difference in Proportion)",
    caption = "Violin trimmed at zero. Box: median, IQR, 1.5\u00d7IQR whiskers. White diamond: mean. Labeled points are upper outliers."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title         = element_text(face = "bold", size = 13),
    plot.subtitle      = element_text(size = 11),
    legend.position    = "none",
    panel.grid.major.x = element_blank(),
    axis.text.x        = element_text(size = 12, face = "bold"),
    plot.margin        = margin(t = 10, r = 40, b = 30, l = 10)
  )

violin_plot

ggsave(
  file.path(out_fig, "violin_baseline_distance.png"),
  plot = violin_plot, width = 12, height = 7, dpi = 300
)
```

------------------------------------------------------------------------

## Domain-Level Distance Aggregation

### Domain-Level Percentage of FTF/Phone Closer to Baseline

```{r domain-aggregation}
closeness_with_topics <- closeness_results %>%
  left_join(topic_mapping, by = "variable") %>%
  mutate(domain = replace_na(as.character(domain), "Other / Uncategorized"))

domain_summary <- closeness_with_topics %>%
  filter(!is.na(dist_to_f2fp), !is.na(dist_to_web)) %>%
  group_by(Domain = domain) %>%
  summarise(
    N               = n(),
    Mean_Dist_FTF   = round(mean(dist_to_f2fp,  na.rm = TRUE), 3),
    Median_Dist_FTF = round(median(dist_to_f2fp, na.rm = TRUE), 3),
    Mean_Dist_Web   = round(mean(dist_to_web,    na.rm = TRUE), 3),
    Median_Dist_Web = round(median(dist_to_web,  na.rm = TRUE), 3),
    Mean_Dist_Full  = round(mean(dist_to_all,    na.rm = TRUE), 3),
    Median_Dist_Full= round(median(dist_to_all,  na.rm = TRUE), 3),
    Pct_FTF_Closer  = round(mean(closer_to == "f2fp", na.rm = TRUE) * 100, 1),
    .groups = "drop"
  ) %>%
  arrange(desc(Pct_FTF_Closer))

gt_domain <- domain_summary %>%
  gt() %>%
  tab_header(
    title    = "Domain-Level Distance from 2016\u20132018 Baseline",
    subtitle = "Mean and median absolute differences by survey mode, pooled 2022\u20132024"
  ) %>%
  cols_label(
    Domain           = "Domain",
    N                = "N vars",
    Mean_Dist_FTF    = "Mean",
    Median_Dist_FTF  = "Median",
    Mean_Dist_Web    = "Mean",
    Median_Dist_Web  = "Median",
    Mean_Dist_Full   = "Mean",
    Median_Dist_Full = "Median",
    Pct_FTF_Closer   = "% FTF Closer"
  ) %>%
  tab_spanner(label = "FTF/Phone Distance", columns = c(Mean_Dist_FTF, Median_Dist_FTF)) %>%
  tab_spanner(label = "Web Distance",       columns = c(Mean_Dist_Web, Median_Dist_Web)) %>%
  tab_spanner(label = "Full Sample Distance", columns = c(Mean_Dist_Full, Median_Dist_Full)) %>%
  fmt_number(
    columns  = c(Mean_Dist_FTF, Median_Dist_FTF, Mean_Dist_Web,
                 Median_Dist_Web, Mean_Dist_Full, Median_Dist_Full),
    decimals = 3
  ) %>%
  fmt_number(columns = Pct_FTF_Closer, decimals = 1) %>%
  tab_source_note(
    "Distance = |p_baseline \u2212 p_mode|. Baseline = pooled 2016\u20132018 FTF sample under WTSSPS."
  ) %>%
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_spanners()
  ) %>%
  tab_options(table.font.size = 11)

gt_domain

gtsave(
  gt_domain,
  filename = file.path(out_tables, "domain_distances.png"),
  expand   = 10
)
```

------------------------------------------------------------------------

## Save Exemplar Data

```{r save-exemplar-data}
# Exemplar classification — saved to CSV for §4 prose verification; no table rendered
exemplar_data_summary <- closeness_results %>%
  left_join(topic_mapping, by = "variable") %>%
  mutate(
    classification = case_when(
      dist_to_f2fp < dist_to_web & dist_to_f2fp < 0.05 ~ "FTF-Preserves",
      dist_to_web  < dist_to_f2fp & dist_to_web  < 0.05 ~ "Web-Preserves",
      abs(dist_to_f2fp - dist_to_web) < 0.02             ~ "Convergent",
      TRUE                                                ~ "Divergent"
    ),
    var_display = gsub("^c_", "", variable) %>% toupper()
  ) %>%
  select(
    Variable       = var_display,
    Domain         = domain,
    Baseline_1618  = est_1618,
    Full_2224      = est_all,
    FTF_2224       = est_f2fp,
    FTF_Distance   = dist_to_f2fp,
    Web_2224       = est_web,
    Web_Distance   = dist_to_web,
    Full_Distance  = dist_to_all,
    FTF_Beats_Full = ftf_beats_full,
    Classification = classification
  ) %>%
  arrange(Classification, Domain, Variable)

write_csv(exemplar_data_summary, file.path(out_data, "exemplar_variables.csv"))
```
