---
title: "GSS Forest Plots — All Variables (Appendix)"
format:
  html:
    toc-depth: 2
editor: visual
---

## Overview

Regenerates all forest plots from `outputs/data/forest_plot_data.csv` (produced by `05_persistence.qmd`):

-   **All-variables forest plot** — 137 variables across 8 domains with 84% confidence intervals by group.
-   **Per-domain forest plots** — Separate plot for each thematic domain.
-   **Racial attitudes domain plot** — Focused visualization of race/ethnicity-related items.

Use this file to edit plot aesthetics without re-running the survey design or confidence interval pipeline.

---

## Setup

```{r setup, message=FALSE, warning=FALSE}
source("00_setup.R")
library(forcats)

# ── Plot-specific aesthetics ───────────────────────────────────────────────────
# group_levels and color_mapping must match MODE_COLORS from 00_setup.R for consistency.

group_levels <- c(
  "2016-2018 (all)",
  "2022-2024 (all)",
  "2022-2024 (f2fp)",
  "2022-2024 (web)"
)

color_mapping <- c(
  "2016-2018 (all)"  = MODE_COLORS["2016-2018"],
  "2022-2024 (all)"  = MODE_COLORS["Population"],
  "2022-2024 (f2fp)" = MODE_COLORS["f2fp"],
  "2022-2024 (web)"  = MODE_COLORS["web"]
)

legend_order <- c(
  "2016-2018 (all)", "2022-2024 (all)",
  "2022-2024 (f2fp)", "2022-2024 (web)"
)

domain_order <- c(
  "Religion",
  "Race & Inequality",
  "Civil Liberties & Government",
  "Crime, Punishment & Firearms",
  "Political & Economic Orientations",
  "Social Attitudes & Morality",
  "Institutional Confidence",
  "Well-being & Life Orientations"
)

# ── Output directories ─────────────────────────────────────────────────────────
out_fig <- "outputs/figures"
out_data <- "outputs/data"
dir.create(file.path(out_fig, "by_domain"), recursive = TRUE, showWarnings = FALSE)
```

---

## Load Data

```{r load-data, message=FALSE, warning=FALSE}
forest_data_path <- file.path(out_data, "forest_plot_data.csv")
if (!file.exists(forest_data_path)) {
  stop("forest_plot_data.csv not found. Run 05_persistence.qmd first.")
}

plot_data_all <- read_csv(forest_data_path, show_col_types = FALSE) %>%
  mutate(
    domain     = factor(domain,     levels = c(domain_order, "Other / Uncategorized")),
    mode_group = factor(mode_group, levels = group_levels),
    variable   = fct_rev(factor(variable))
  )
```

---

## Helper: Plot Theme

```{r theme-helper}
# Centralised theme so tweaks propagate to all plots
forest_theme <- function(base_size = 11) {
  theme_bw(base_size = base_size) +
    theme(
      plot.title        = element_text(hjust = 0.5, face = "bold", size = base_size + 3),
      plot.subtitle     = element_text(hjust = 0.5, size = base_size),
      legend.position   = "bottom",
      legend.text       = element_text(size = base_size - 2),
      legend.title      = element_text(face = "bold"),
      strip.placement   = "outside",
      strip.background  = element_blank(),
      strip.text.y.left = element_text(angle = 0, face = "bold", hjust = 1),
      panel.grid.minor  = element_blank()
    )
}
```

---

## Violin Plot: Distance from Baseline

```{r load-closeness, message=FALSE, warning=FALSE}
closeness_path <- file.path(out_data, "closeness_by_variable.csv")
if (!file.exists(closeness_path)) {
  stop("closeness_by_variable.csv not found. Run 05_persistence.qmd first.")
}

closeness_results <- read_csv(closeness_path, show_col_types = FALSE)
```

```{r distance-plots, fig.width=12, fig.height=7}
closeness_long <- closeness_results %>%
  filter(!is.na(dist_to_f2fp) & !is.na(dist_to_web) & !is.na(dist_to_all)) %>%
  pivot_longer(
    cols      = c(dist_to_f2fp, dist_to_all, dist_to_web),
    names_to  = "mode",
    values_to = "distance"
  ) %>%
  mutate(
    mode_label = case_when(
      mode == "dist_to_f2fp" ~ "FTF/Phone",
      mode == "dist_to_all"  ~ "Full Sample\n(native weights)",
      mode == "dist_to_web"  ~ "Web"
    ),
    mode_label = factor(mode_label,
                        levels = c("FTF/Phone", "Full Sample\n(native weights)", "Web"))
  )

distance_stats <- closeness_long %>%
  group_by(mode_label) %>%
  summarise(
    median = median(distance, na.rm = TRUE),
    mean   = mean(distance, na.rm = TRUE),
    n      = n(),
    .groups = "drop"
  )

violin_fills <- c(
  "FTF/Phone"                      = "#D95F02",
  "Full Sample\n(native weights)"  = "#888888",
  "Web"                            = "#1B9E77"
)

# Identify upper outliers (beyond 1.5 * IQR) per mode group
outlier_data <- closeness_long %>%
  group_by(mode_label) %>%
  mutate(
    q1         = quantile(distance, 0.25, na.rm = TRUE),
    q3         = quantile(distance, 0.75, na.rm = TRUE),
    iqr        = q3 - q1,
    is_outlier = distance > (q3 + 1.5 * iqr)
  ) %>%
  filter(is_outlier) %>%
  mutate(
    var_label = toupper(gsub("^c_", "", variable)),
    hjust_val = if_else(mode_label == "FTF/Phone", 1.15, -0.15),
    nudge_x   = if_else(mode_label == "FTF/Phone", -0.07, 0.07)
  ) %>%
  ungroup()

violin_plot <- ggplot(closeness_long,
                      aes(x = mode_label, y = distance, fill = mode_label)) +
  geom_violin(alpha = 0.55, trim = TRUE) +
  geom_boxplot(width = 0.15, alpha = 0.85, outlier.shape = NA, color = "gray20") +
  geom_point(data = outlier_data,
             aes(x = mode_label, y = distance, color = mode_label),
             size = 2.5, inherit.aes = FALSE) +
  geom_text(data = outlier_data,
            aes(x     = as.numeric(mode_label) + nudge_x,
                y     = distance,
                label = var_label,
                hjust = hjust_val),
            size = 4, color = "gray25", inherit.aes = FALSE) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 4,
               fill = "white", color = "black") +
  scale_fill_manual(values  = violin_fills, name = NULL) +
  scale_color_manual(values = violin_fills, guide = "none") +
  # Embed Md/M stats directly in x-axis tick labels
  scale_x_discrete(
    labels = setNames(
      sprintf("%s\nMd = %.3f  M = %.3f",
              c("FTF/Phone", "Full Sample\n(native weights)", "Web"),
              distance_stats$median,
              distance_stats$mean),
      levels(closeness_long$mode_label)
    )
  ) +
  scale_y_continuous(
    labels = scales::number_format(accuracy = 0.01),
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.08))
  ) +
  labs(
    title    = "Distribution of Distance from 2016\u20132018 Baseline",
    subtitle = "Absolute difference |p_baseline \u2212 p_mode| across all variables",
    x        = NULL,
    y        = "Distance from Baseline (Absolute Difference in Proportion)",
    caption  = "Violin trimmed at zero. Box: median, IQR, 1.5\u00d7IQR whiskers. White diamond: mean. Labeled points are upper outliers."
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title         = element_text(face = "bold", size = 20),
    plot.subtitle      = element_text(size = 16),
    plot.caption       = element_text(size = 13),
    legend.position    = "none",
    panel.grid.major.x = element_blank(),
    axis.text.x        = element_text(size = 15, face = "bold", lineheight = 1.4),
    axis.text.y        = element_text(size = 14),
    axis.title.y       = element_text(size = 15),
    plot.margin        = margin(t = 10, r = 60, b = 10, l = 10)
  )

violin_plot

ggsave(
  file.path(out_fig, "violin_baseline_distance.png"),
  plot = violin_plot, width = 12, height = 7, dpi = 300
)
```

---

## All-Variables Forest Plot (Appendix)

```{r forest-all, fig.width=12, fig.height=20}
forest_all <- ggplot(plot_data_all,
                     aes(x = estimate, y = variable, color = mode_group)) +
  geom_point(position = position_dodge(width = 0.7), size = 2) +
  geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper),
                 position = position_dodge(width = 0.7),
                 height = 0, linewidth = 0.7) +
  facet_grid(domain ~ ., scales = "free_y", space = "free_y", switch = "y") +
  scale_color_manual(values = color_mapping, breaks = legend_order, name = "Group") +
  guides(color = guide_legend(nrow = 2, byrow = TRUE)) +
  labs(
    title    = "84% Confidence Intervals by Domain and Group",
    subtitle = "GSS 2016–18 baseline vs. 2022–24, by mode",
    x        = "Estimated Proportion",
    y        = NULL
  ) +
  coord_cartesian(xlim = c(0, 1)) +
  forest_theme()

forest_all

plot_height <- max(8, n_distinct(plot_data_all$variable) * 0.5)
ggsave(
  file.path(out_fig, "forest_all_variables.png"),
  plot = forest_all, width = 12, height = plot_height,
  units = "in", dpi = 300, limitsize = FALSE
)
```

---

## Per-Domain Forest Plots

```{r forest-by-domain, message=FALSE}
safe_name <- function(x) {
  tolower(x) %>%
    str_replace_all("[^a-z0-9]+", "_") %>%
    str_remove("_$")
}

plot_domain <- function(domain_name, data) {
  domain_data <- data %>% filter(as.character(domain) == domain_name)
  n_vars      <- n_distinct(domain_data$variable)

  p <- ggplot(domain_data,
              aes(x = estimate, y = variable, color = mode_group)) +
    geom_point(position = position_dodge(width = 0.7), size = 2.5) +
    geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper),
                   position = position_dodge(width = 0.7),
                   height = 0, linewidth = 0.8) +
    scale_color_manual(values = color_mapping, breaks = legend_order, name = "Group") +
    guides(color = guide_legend(nrow = 2, byrow = TRUE)) +
    labs(
      title    = domain_name,
      subtitle = "84% CIs — GSS 2016–18 baseline vs. 2022–24, by mode",
      x        = "Estimated Proportion",
      y        = NULL
    ) +
    coord_cartesian(xlim = c(0, 1)) +
    forest_theme()

  list(plot = p, height = max(4, n_vars * 0.45))
}

out_fig_domains <- file.path(out_fig, "by_domain")
domains_present <- levels(droplevels(plot_data_all$domain))

for (dom in domains_present) {
  result <- plot_domain(dom, plot_data_all)
  fname  <- paste0("forest_", safe_name(dom), ".png")
  ggsave(
    file.path(out_fig_domains, fname),
    plot = result$plot, width = 10, height = result$height,
    units = "in", dpi = 300, limitsize = FALSE
  )
}
```

---

## Racial Attitudes Domain

```{r forest-racial, fig.width=12}
racial_topics <- unique(plot_data_all$domain)[
  str_detect(tolower(as.character(unique(plot_data_all$domain))), "racial|race")
]

plot_data_racial <- plot_data_all %>% filter(domain %in% racial_topics)

if (nrow(plot_data_racial) > 0) {
  forest_racial <- ggplot(plot_data_racial,
                          aes(x = estimate, y = variable, color = mode_group)) +
    geom_point(position = position_dodge(width = 0.7), size = 2.5) +
    geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper),
                   position = position_dodge(width = 0.7),
                   height = 0, linewidth = 0.8) +
    facet_grid(domain ~ ., scales = "free_y", space = "free_y", switch = "y") +
    scale_color_manual(values = color_mapping, breaks = legend_order, name = "Group") +
    guides(color = guide_legend(nrow = 2, byrow = TRUE)) +
    labs(
      title    = "84% Confidence Intervals — Racial Attitudes Domain",
      subtitle = "GSS 2016–18 baseline vs. 2022–24, by mode",
      x        = "Estimated Proportion",
      y        = NULL,
      caption  = paste0(
        "84% non-overlapping CIs ≈ p < .05 pairwise difference.\n",
        "FTF/Phone (orange); Web (teal); Full sample (black); Baseline 2016–18 (gray)."
      )
    ) +
    coord_cartesian(xlim = c(0, 1)) +
    forest_theme()

  forest_racial

  plot_height_racial <- max(6, n_distinct(plot_data_racial$variable) * 0.5)
  ggsave(
    file.path(out_fig, "forest_racial_attitudes.png"),
    plot = forest_racial, width = 12, height = plot_height_racial,
    units = "in", dpi = 300, limitsize = FALSE
  )
}
```
